---
title: "Lab 02: Basic Data Visualization"
author: "Your Name Here"
date: "Today's Date Here"
format: html
---

```{r }
#| label: data-and-packages
#| warning: false
#| message: false

library(tidyverse)

# Load in the dataset in the R chunk here! Note that you MUST load in the dataset
# in the chunk in order for other R chunks to recognize the dataset, even if
# you've loaded it in the Console.
cdc <- read.csv("https://karamccor.github.io/b6002/labs/data/cdc_cleaned.csv")
```

### Exercise 1

Write your narrative to Exercise 1 here.

### Exercise 2

```{r }
#| label: ex-2

# Use this R chunk to create your visualization for Exercise 2.
```

Write your narrative to Exercise 2 here.

### Exercise 3

```{r }
#| label: ex-3

# Use this R chunk to create your visualization for Exercise 3.
```

Write your narrative to Exercise 3 here.

### Exercise 4

```{r }
#| label: ex-4

# Use this R chunk to create your visualization for Exercise 4.
```

Write your narrative to Exercise 4 here.

### Exercise 5

```{r }
#| label: ex-5

# Use this R chunk to create your visualization for Exercise 5.
```

Write your narrative to Exercise 5 here.

### Exercise 6

```{r }
#| label: ex-6
# Use this R chunk to create your visualization for Exercise 6.
```

Write your narrative to Exercise 6 here.




**Submission:** Remember to render your lab into an .html file and then save it as a .pdf in a web browser. You must turn in a .pdf file corresponding to the Quarto template to Gradescope in order to receive credit for the labs and homework assignments. You must assign each problem to a page in your PDF. Assign the Formatting portion to your page 1. 

