---
title: "Lecture 15: Logistic regression"
subtitle: "BIOS 600 - Summer Session II 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format: html
---

## Announcements {.smaller}

- HW 15 due **tonight** 

- Hand in Application Exercise 3 if you haven't already, and you can resubmit if you lost points on your first submission.

- Make an appointment to go over your Exam 2 for points back!

- Fill out your course evaluation by tomorrow please; if the response rate gets to 80% everyone in the class will have one point added to their final exam score.

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(tidymodels)
library(kableExtra)
library(MASS)
library(randomForestSRC) # for pbc data
data(pbc, package = "randomForestSRC")
cdc <- read.csv("https://karamccor.github.io/b6002/labs/data/cdc_cleaned.csv")

cdc <- cdc |>
  mutate(over65_bin = as.numeric(Over65 > 14.834))
```

## Predictions with two categorical variables

Suppose we use our `cdc` dataset and create a new variable called `over65_bin`, which equals 1 if a state has an above-average population over 65 and 0 otherwise. We might want to run the following model:

$$CVDeathRate_i = \beta_0 + \beta_1(HDI == Middle)_i + \beta_2(HDI == TopTen)_i$$
$$+ \beta_3(over65\_bin == 1)_i.$$

```{r}
#| echo: FALSE


fit_death <- lm(CVDeathRate ~ HDI + over65_bin, 
                data = cdc)

tidy(fit_death) |>
  kable(digits = 2)

```

Our fitted model is then

$$\widehat{CVDeathRate}_i = 209.54 - 42.14(HDI == Middle)_i - 62.50(HDI == TopTen)_i$$
$$- 6.56(over65\_bin == 1)_i.$$

The graph of this would look like this:

```{r}
#| echo: FALSE


ggplot(cdc) +
  geom_hline(yintercept = 209.54, color = "coral1") +
  annotate("text", label = "Bottom 10 HDI, Below-average 65+ Population", 
           x = 20, y = 213, size = 2.5) +
  geom_hline(yintercept = 209.54-6.56, color = "coral3") +
  annotate("text", label = "Bottom 10 HDI, Above-average 65+ Population", 
           x = 20, y = 206, size = 2.5) +
  geom_hline(yintercept = 209.54 - 42.14, color = "darkslategray2") +
  annotate("text", label = "Middle HDI, Below-average 65+ Population", 
           x = 20, y = 170, size = 2.5) +
  geom_hline(yintercept = 209.54 - 42.14 - 6.56, color = "darkslategray4") +
  annotate("text", label = "Middle 10 HDI, Above-average 65+ Population", 
           x = 20, y = 163, size = 2.5) +
  geom_hline(yintercept = 209.54 - 62.50, color = "darkorchid2") +
  annotate("text", label = "Top 10 HDI, Below-average 65+ Population", 
           x = 20, y = 150, size = 2.5) +
  geom_hline(yintercept = 209.54 - 62.50 - 6.56, color = "darkorchid4") +
  annotate("text", label = "Top 10 HDI, Above-average 65+ Population", 
           x = 20, y = 143, size = 2.5) +
  coord_cartesian(ylim = c(115, 215), xlim = c(15, 22)) +
  theme_light()


```

Now suppose we want to include an interaction term between HDI category and `over65_bin`. Now our model looks like:

$$CVDeathRate_i = \beta_0 + \beta_1(HDI == Middle)_i + \beta_2(HDI == TopTen)_i $$
$$+ \beta_3(over65\_bin == 1)_i + \beta_4(HDI == Middle)_i(over65\_bin == 1)_i $$
$$+ \beta_5(HDI == TopTen)_i(over65\_bin== 1)_i.$$

```{r}
#| echo: FALSE

fit_death_int <- lm(CVDeathRate ~ HDI + over65_bin + HDI*over65_bin, 
                data = cdc)

tidy(fit_death_int) |>
  kable(digits = 2)

```

Our fitted model is then

$$\widehat{CVDeathRate}_i = 221.10 - 56.18(HDI == Middle)_i - 75.55(HDI == TopTen)_i$$
$$- 25.82(over65\_bin == 1)_i + 24.23(HDI == Middle)_i(over65\_bin == 1)_i$$
$$ + 22.99(HDI == TopTen)_i(over65\_bin == 1)_i.$$

The graph of this would look like this:

```{r}
#| echo: FALSE

ggplot(cdc) +
  geom_hline(yintercept = 221.10, color = "coral1") +
  annotate("text", label = "Bottom 10 HDI, Below-average 65+ Population", 
           x = 20, y = 225, size = 2.5) +
  geom_hline(yintercept = 221.10 - 25.82, color = "coral3") +
  annotate("text", label = "Bottom 10 HDI, Above-average 65+ Population", 
           x = 20, y = 199, size = 2.5) +
  geom_hline(yintercept = 221.10 - 56.18, color = "darkslategray2") +
  annotate("text", label = "Middle HDI, Below-average 65+ Population", 
           x = 20, y = 168, size = 2.5) +
  geom_hline(yintercept = 221.10 - 56.18 - 25.82 + 24.23, color = "darkslategray4") +
  annotate("text", label = "Middle 10 HDI, Above-average 65+ Population", 
           x = 20, y = 160, size = 2.5) +
  geom_hline(yintercept = 221.10 - 75.55, color = "darkorchid2") +
  annotate("text", label = "Top 10 HDI, Below-average 65+ Population", 
           x = 20, y = 149, size = 2.5) +
  geom_hline(yintercept = 221.10 - 75.55 - 25.82 + 22.99, color = "darkorchid4") +
  annotate("text", label = "Top 10 HDI, Above-average 65+ Population", 
           x = 20, y = 140, size = 2.5) +
  coord_cartesian(ylim = c(100, 250), xlim = c(15, 22)) +
  theme_light()


```
