---
title: "Lecture 14: Non-Parametric tests"
subtitle: "BIOS 600 - Spring 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Announcements

-   HW 6 due Thursday **March 12th** at 11:59pm. 

-   Lab this week has shifted to TA office hours, Lab 7 will not be due until after Exam 2

-   Encourage you all to attend follow up on questions about HW 6 and/or understanding any questions remaining on previous homeworks and labs

-   Professor Howard is out of the office and **will not** have office hours

## Very optional reading

-   P&G Chapters 13

-   OI: Sections 8.1

## Overview

- Nonparametric methods

- Sign test, Wilcoxon signed-rank test, Wilcoxon rank-sum test

- Krustal-Wallis test

## Why nonparametric methods? {.smaller}

-   For the methods we have studied so far, we have assumed the populations from which the data were drawn were either normally distributed or approximately so. This is a necessary property for the tests to be valid.

-   Because the distributional form is assumed known, with only values of $\mu$ and $\sigma$ unknown, these methods are known as **parametric** methods (e.g. t-test, ANOVA).

-   **Nonparametric methods** make fewer assumptions about the nature of the underlying distributions and may be appropriate when some assumptions of parametric methods are not satisfied (small sample size).

## Sign test

-   The **sign test** is a nonparametric alternative to the paired t-test. It does not assume normality but just requires that observations are independent.

-   The null hypothesis of the sign test is that the median difference among pairs in the underlying population is 0.

## Sign test: cold data {.smaller}

-   A medical researcher claims that a new vaccine will decrease the number of colds in adults.

-   You randomly select 14 adults and record the number of colds each has in a one-year period.

-   After giving the vaccine to each adult, you again record the number of colds each has in a one-year period.

-   At $\alpha = 0.05$, do the data support the researcher’s claim?

| \# Colds |     |     |     |     |     |     |     |     |     |     |     |     |     |     |
|----------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| Before   | 3   | 4   | 2   | 1   | 3   | 6   | 4   | 5   | 2   | 0   | 2   | 5   | 3   | 3   |
| After    | 2   | 1   | 0   | 3   | 1   | 3   | 3   | 2   | 2   | 2   | 2   | 3   | 4   | 3   |

## Sign test

The sign test has just a few simple steps:

-   Take the difference of each pair of observations (e.g., before-after)

-   Record the sign of each difference (+, --, or 0)

-   Count the number of + signs

-   Test using the binomial distribution with $n =$ number of pairs with nonzero differences and $p=0.5$, as we would expect the **same number of positive and negative signs** under $H_0$.

## Sign test {.smaller}

The differences do NOT look very normal and we have a relatively small sample size. So, we cannot use previous methods to solve this question (and must rely on the sign test)!

\footnotesize

| \# Colds |     |     |     |     |     |     |     |     |     |     |     |     |     |     |
|----------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| Before   | 3   | 4   | 2   | 1   | 3   | 6   | 4   | 5   | 2   | 0   | 2   | 5   | 3   | 3   |
| After    | 2   | 1   | 0   | 3   | 1   | 3   | 3   | 2   | 2   | 2   | 3   | 4   | 3   | 2   |
| Diff.    | 1   | 3   | 2   | -2  | 2   | 3   | 1   | 3   | 0   | -2  | -1  | 1   | 0   | 1   |
| Sign     | \+  | \+  | \+  | --  | \+  | \+  | \+  | \+  | 0   | --  | --  | \+  | 0   | \+  |

::: table-striped
12 non-zero differences, with 9 + signs
:::

## Sign test for cold data

$H_0$: Median difference in colds = 0 (i.e. same number of positive and negative signs)

$H_A$: Median difference in colds $\neq$ 0 (i.e. different number of positive and negative signs)

-   We have two possible outcomes (positive **or** negative signs)

-   With two outcomes possible, we have a binomial distribution, with $p$ = Probability of success (defined by number of positive signs)

## Sign test for cold data

$H_0$: Median difference in colds = 0 (i.e. same number of positive and negative signs)

$H_A$: Median difference in colds $\neq$ 0 (i.e. different number of positive and negative signs)

-   Remember, we had 9 successes out of 12 trials.

-   What is the probability we see 9 successes out of 12 trials in a binomial distribution (or more extreme, *in either direction*), if the null hypothesis that $p=0.5$ is true?

## Conducting the sign test

We can calculate, for $X\sim Binom(12,0.5)$,
<br>
$P(X=9) + P(X=10) + P(X=11) + P(X=12)$
<br>
**Remember $P(X\geq 9) = 1-P(X<8)$**

and then double it (for the two-sided test).

```{r echo = T}
2*(1-pbinom(8, 12, 0.5))
```

## Conducting the sign test

-   **Conclusion**: With p = .145, we fail to reject $H_0$ and hence conclude that there is not enough evidence to suggest there is a difference in the median number of colds in the years before and after vaccine administration.

## Wilcoxon signed-rank test {.smaller}

-   The sign test is appealing because it avoids all distributional assumptions. However, it ignores the magnitude of the differences.

-   If we are willing to assume that the differences are symmetrically distributed around the median, we can incorporate the magnitude of the differences and gain considerable power using the **Wilcoxon Signed-Rank Test**.

-   The Wilcoxon signed-rank test is an alternative to the paired t-test. The $H_0$ is that the median difference in the underlying population is 0.

## Rank-based methods

-   Many non-parametric methods have a similar flavor to parametric methods but are computed on the ranks instead of the observed data.

-   The rank of an observation, among a set of observations, is its position when the observations are ordered from smallest to largest. The smallest observation has rank 1, next smallest has rank 2, and so forth.

-   If observations are tied, the rank assigned to each is the average of the ranks appropriate to the equal numbers.

## Why ranks?

Using the ranks provides robustness against outliers:

-   Weights: 175, 169, 190

    -   Ranks: 2, 1, 3

-   Weights: 175, 169, 490

    -   Ranks: 2, 1, 3 (robust to outlier)

-   Weights: 175, 175, 190

    -   Ranks: 1.5, 1.5, 3 (take sum of ranks and divide by number tied)

## Wilcoxon signed-rank test

Like the sign test, the **Wilcoxon signed-rank test** is for paired differences and has just a few simple steps:

-   Calculate the difference for each pair of observations

-   Rank the absolute values of these differences from smallest to largest (drop 0's; assign an average rank to ties); let $n$ be the number of non-zero differences

-   Assign each rank a + or -- depending on the original sign of the difference

-   Add up all positive ranks; add up all negative ranks; let $T$ be equal to the smaller sum

## Wilcoxon signed-rank test

Calculate the z-score

::: poll
$$z_T = \frac{T - \frac{n(n+1)}{4}}{\sqrt{\frac{n(n+1)(2n+1)}{24}}}$$
:::

and obtain a p-value from the normal distribution.

Don't do this by hand -- we will use R. 

## Example: middle ear effusion {.smaller}

-   A common symptom of otitis media in young children is the prolonged presence of fluid in the middle ear, known as middle-ear effusion. The presence of fluid may result in temporary hearing loss and interfere with normal learning skills in the first 2 years of life.

-   One hypothesis is that babies who are breastfed for at least one month build some immunity and have less prolonged effusion than their bottle-fed counterparts.

-   A small study of 14 babies is set up, with babies matched one-to-one by age, sex, socioeconomic status, and health condition. One member of the pair is a breastfed baby, and the other member is a bottle-fed baby. The outcome is the duration (days) of middle-ear effusion after the first episode of otitis media.

## Example: middle ear effusion {.smaller}

Duration of Effusion for Each Pair (Days):

| Pair | 1   | 2   | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  | 11  | 12  | 13  | 14  |
|-------------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| Breast   | 26  | 3   | 12  | 28  | 7   | 39  | 12  | 30  | 7   | 15  | 65  | 10  | 19  | 11  |
| Bottle  | 18  | 7   | 6   | 33  | 7   | 57  | 29  | 28  | 8   | 27  | 78  | 17  | 16  | 35  |

::: {.callout-note appearance="simple"}
## Question

What is $H_0$?
:::

## Example: middle ear effusion {.smaller}

Duration of Effusion for Each Pair (Days):

\tiny

|                 | 1   | 2   | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  | 11  | 12  | 13  | 14  |
|-------------|-----|-----|-----|-----|-----|--------|--------|------|-----|---------|------|------|-----|--------|
| Breast   | 26  | 3   | 12  | 28  | 7   | 39  | 12  | 30  | 7   | 15  | 65  | 10  | 19  | 11  |
| Bottle  | 18  | 7   | 6   | 33  | 7   | 57  | 29  | 28  | 8   | 27  | 78  | 17  | 16  | 35  |
| Diff  | 8   | -4  | 6   | -5  | 0   | -18 | -17 | 2   | -1  | -12 | -13 | -7  | 3   | -24 |
| Abs(Diff)  | 8   | 4   | 6   | 5   | 0   | 18  | 17  | 2   | 1   | 12  | 13  | 7   | 3   | 24  |
| Rank        | 8   | 4   | 6   | 5   | NA  | 12  | 11  | 2   | 1   | 9   | 10  | 7   | 3   | 13  |
| Signed Rank | 8   | -4  | 6   | -5  | NA  | -12 | -11 | 2   | -1  | -9  | -10 | -7  | 3   | -13 |


\normalsize

13 non-zero ranks

-   Sum of positive ranks is smaller than sum of negative ranks

-   Sum of positive ranks = 8 + 6 + 2 + 3 = 19 = $T$

## Example: middle ear effusion {.smaller}

Duration of Effusion for Each Pair (Days):

\tiny

|                 | 1   | 2   | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  | 11  | 12  | 13  | 14  |
|-------------|-----|-----|-----|-----|-----|--------|--------|------|-----|---------|------|------|-----|--------|
| Breast   | 26  | 3   | 12  | 28  | 7   | 39  | 12  | 30  | 7   | 15  | 65  | 10  | 19  | 11  |
| Bottle  | 18  | 7   | 6   | 33  | 7   | 57  | 29  | 28  | 8   | 27  | 78  | 17  | 16  | 35  |
| Diff  | 8   | -4  | 6   | -5  | 0   | -18 | -17 | 2   | -1  | -12 | -13 | -7  | 3   | -24 |
| Abs(Diff)  | 8   | 4   | 6   | 5   | 0   | 18  | 17  | 2   | 1   | 12  | 13  | 7   | 3   | 24  |
| Rank        | 8   | 4   | 6   | 5   | NA  | 12  | 11  | 2   | 1   | 9   | 10  | 7   | 3   | 13  |
| Signed Rank | 8   | -4  | 6   | -5  | NA  | -12 | -11 | 2   | -1  | -9  | -10 | -7  | 3   | -13 |


\normalsize

13 non-zero ranks

-   $z_T = \frac{19-\frac{13(14)}{4}}{\sqrt{\frac{13(14)(27)}{24}}} = \frac{19-45.5}{3} = -1.85$ $\rightarrow$ Now, let's calculate our p-value.

## Calculating the p-value

```{r}
#| echo: true
zt <- (19-(13*14/4))/sqrt(13*14*27/24)
zt
pnorm(zt)*2 # two-sided test
```

- **Conclusion**: With $p > 0.05$, we fail to reject the null hypothesis. We do not have sufficient evidence to suggest that there is a difference in the median of the paired differences in effusion duration between being breast and bottle-fed.


## Wilcoxon rank sum test

Not to be confused with the Wilcoxon Signed-Rank Test! 

The **Wilcoxon Rank Sum Test**:

-   Is the non-parametric counterpart of the two-sample t-test

-   Assumes two samples **are independent** but does not require normality or equal variance of groups

-   Assumes two distributions have "roughly the same shape"

-   Also called the **Mann-Whitney U test** or **Wilcoxon-Mann-Whitney test**

## Mann-Whitney U test

-   The Mann-Whitney U-test is an extension of the Wilcoxon signed rank test for **independent samples**.

-   Instead of taking differences, the observations are lumped together and ranked, with then the sum of the ranks calculated by group.

-   We'll use software to perform this test.

## Example: Exercise capacity and coronary artery disease (CAD) {.smaller}

-   Consider data from a two-group study of exercise capacity.

-   Two groups of men, one with diagnosed three-vessel coronary artery disease (3VD), and the other group of men with suspected disease (SD) in one or more vessels.

-   The total time participants could exercise on a treadmill set to increase in speed and slope according to a set schedule, is below:

-   3VD times: 864, 636, 638, 708, 786, 600, 1320, 750, 594, 750

-   SD times: 1014, 684, 810, 990, 840, 978, 1002, 1110

-   3VD median: 729

-   SD median: 984

## Example: Exercise capacity and CAD {.smaller}



| Value | Rank | Group | Value | Rank | Group |
|-------|------|-------|-------|------|-------|
| 594   | 1    | 3VD   | 810   | 10   | SD    |
| 600   | 2    | 3VD   | 840   | 11   | SD    |
| 636   | 3    | 3VD   | 864   | 12   | 3VD   |
| 638   | 4    | 3VD   | 978   | 13   | SD    |
| 684   | 5    | SD    | 990   | 14   | SD    |
| 708   | 6    | 3VD   | 1002  | 15   | SD    |
| 750   | 7.5  | 3VD   | 1014  | 16   | SD    |
| 750   | 7.5  | 3VD   | 1110  | 17   | SD    |
| 786   | 9    | 3VD   | 1320  | 18   | 3VD   |



-   Sum of ranks in SD group, $n=8$: 101

-   Sum of ranks in 3VD group, $n=10$: 70

## Example

-   The sum of the ranks in the SD group with $n=8$ was 101, and the 3VD group had 10 subjects with sum of ranks 70. Is this enough evidence to reject the null hypothesis? 

## Mann-Whitney U Test Statistic

| Group | n  | Sum of Ranks |
|:------|:--:|:-------------:|
| SD    | 8  | 101           |
| 3VD   | 10 | 70            |

**Step 1** Compute U-statistics.

$$U_1 = R_1 - \frac{n_1(n_1+1)}{2}, \quad U_2 = R_2 - \frac{n_2(n_2+1)}{2}$$

where $R_1$ and $R_2$ are the sums of ranks in groups 1 and 2, respectively, and $n_1$ and $n_2$ are group sample sizes.

## U-statistic

- When both sample sizes are $\geq 8$, the sampling distribution of U is approximately normal with mean and standard deviation:

$$\text{mean}_U = \frac{U_1 + U_2}{2} = \frac{n_1 n_2}{2}, \quad \text{sd}_U = \sqrt{\frac{n_1 n_2 (n_1 + n_2 + 1)}{12}}$$

- In other words, $U_1 + U_2 = n_1 \cdot n_2$, so once you know one, you automatically know the other. 

**No need to memorize this!**


## In R

```{r}
#| echo: true
# Given data
n1 <- 8; n2 <- 10
R1 <- 101; R2 <- 70

U1 <- R1 - n1 * (n1 + 1) / 2
U2 <- R2 - n2 * (n2 + 1) / 2
U1; U2
```

## In R

```{r}
#| echo: true
mean_U <- n1 * n2 / 2
sd_U <- sqrt(n1 * n2 * (n1 + n2 + 1) / 12)
z <- (U1 - mean_U) / sd_U
p_value <- 2 * (1 - pnorm(z))

z; p_value
```

-   As it turns out, we reject $H_0$ and conclude the two groups do not have the same median. The median exercise time in the SD group is higher compared to the 3VD group.

## Why do we use $U_1$ when computing $z$? {.smaller}

- When approximating the test statistic using the normal distribution:

$$z = \frac{U - \text{mean}_U}{\text{sd}_U}$$

you can use either $U_1$ or $U_2$ - they'll lead to the same p-value once you take into account the direction of the test. 

- By convention, we usually use $U_1$, the statistic associated with the first group. 

- Using $U_1$ or $U_2$ would yield the same absolute z-score, since $U_1$ and $U_2$ are complements. 

## Check

- Using $U_2$: 

```{r}
#| echo: true
z <- (U2 - mean_U) / sd_U
p_value <- 2 * (pnorm(z))

z; p_value
```
We see that our $z$ value is simply the negative of the previous $z$ value. 
Using a two-sided test, we get the same p-value.

## Note

::: callout-tip

## Alternative way of calculating p-value

- Note that we can also calculate our p-value using:

```{r}
#| echo: true
2 * pnorm(-abs(z))
```

- Why can we do this? Can you draw a picture to explain why these two methods yield the same result? (Look up `?abs` in the Console)

- Why might this code be **more generalizable** compared to the previous code to calculate the p-value?

:::

## Kruskal-Wallis test

-   The **Kruskal-Wallis Test** is the nonparametric version of ANOVA, generalizing the Wilcoxon rank sum test to more than 2 groups.

**Kruskal-Wallis test for pet data**

-   Consider the study of pulse rate from the ANOVA lecture.

-   Remember that pulse rates were not necessarily normally distributed within groups.

## Kruskal-Wallis test for pet data

```{r}
library(tidyverse)
# Set seed for reproducibility
set.seed(123)

# Simulate data: 30 people per group
puppies <- rnorm(30, mean=78, sd=7)  # lower pulse for those with puppies
neither <- rnorm(30, mean=82, sd=5)    # pulse for those alone
friends <- rnorm(30, mean=85, sd=5)  # pulse for those chatting with a friend

# Combine into a data frame
pulse_data <- data.frame(
  pulse = c(friends, neither, puppies),
  group = factor(rep(c("Friend", "Neither", "Puppies"), each=30))
)

# Calculate the ranks of the continuous variable (value), group by group, and then summarize
ranked_data <- pulse_data %>%
  mutate(rank = rank(pulse)) %>%         # Rank the continuous variable
  group_by(group) %>%                    # Group by the categorical variable (group)
  summarize(sum_of_ranks = sum(rank))    # Summarize the sum of ranks within each group
```

-   Calculating the sum of the ranks, we find that in the pet group, out of $n=30$, the sum of the ranks was 863; out of the $n=30$ in the friend group, the sum of the ranks was 1797, and out of the $n=30$ where neither were present, the sum of the ranks was 1435. Is this enough evidence to reject the null hypothesis? (Ask software).

```{r echo = T}
kruskal.test(pulse ~ group, data = pulse_data)
```

The p-value is \< 0.0001. What may we conclude?

## Pairwise comparisons: pet vs. friend

-   Again, we can step-down and look at pairwise differences. In this case, we would use the Wilcoxon Rank-Sum test (why?).

## When to Use Kruskal–Wallis vs. ANOVA {.smaller}

\footnotesize

| Feature | One-Way ANOVA | Kruskal–Wallis Test |
|:---------|:--------------|:--------------------|
| **Purpose** | Tests whether the **means** of 3+ groups differ | Tests whether the **distributions** (typically medians) differ |
| **Data type** | Continuous, roughly normal | Ordinal or continuous, not normal |
| **Assumptions** | 1. Normality within groups <br> 2. Equal variances (homoscedasticity) <br> 3. Independent samples | 1. Independent samples <br> 2. Same-shaped distributions (for median interpretation) |
| **Basis of test** | Uses **group means** and **within-group variance** | Uses **ranks** of data instead of raw values |
| **Example use** | Compare mean blood pressure across 3 treatment groups | Compare pain severity scores (on a 0–10 scale) across 3 treatment groups when scores are skewed |

## Quick Intuition


- Use **ANOVA** if your data meet **parametric assumptions** (normal, equal variances).  

- Use **Kruskal–Wallis** if your data are **skewed**, **ordinal**, or have **outliers** that violate ANOVA’s assumptions.

## Why not always go nonparametric?

-   Non-parametric methods are desirable because they do not require as many restrictive assumptions as parametric ones.

-   However, this flexibility comes at a price – if the assumptions underlying a parametric test are satisfied, the nonparametric test is **less powerful** than the comparable parametric technique. We'll talk about statistical power next class.

## Next Class

- Power and sample size calculation (incredibly useful!)