---
title: "Lecture 8: The central limit theorem"
subtitle: "BIOS 600 - Spring 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Announcements

- Tuesday 2/17: Exam 01 in-class. All you need is something to write with. 

- Thursday 2/12 class will be a review session. 

- Last names starts with A-K - in McGavran-Greenberg Hall 2301

- Last names start with L-Z you will be in this classroom

## Announcements

- HW 3 due Thursday 2/12 at 11:59pm. 

- Lab 4 due Friday 2/13 at 11:59 pm

## Announcements

- Swing by lab session if you need help on Lab 4, HW 3, or exam review.

- **Reminder** Free tutoring is available!!! 


## Application Exercise from Last week

Suppose 5 people receive a new vaccine. From clinical trial data, the probability that any individual experiences a mild side effect (e.g., sore arm, mild fever) is 0.2.  

Let $X$ be the number of people (out of 5) who experience a mild side effect.  

Then $X \sim \text{Binomial}(n=5, p=0.2)$.

## Exercise 1

What is the probability that exactly 2 out of 5 people experience a mild side effect? Use the `dbinom` function to compute your answer and provide a 1-sentence conclusion.

```{r, echo=TRUE}
dbinom(2, size=5, prob=0.2)
```

## Exercise 2

What is the probability that at least 1 out of 5 people experiences a mild side effect?
(Hint: Use the complement rule: $P(X \geq 1)=1−P(X=0)$). Use the `dbinom` function to compute your answer and provide a 1-sentence conclusion.

```{r, echo=TRUE}
p0 <- dbinom(0, size=5, prob=0.2)
p0
1-p0
```

## Hospital Visits: Poisson Example

Suppose that in a hospital emergency department, the average number of asthma-related visits per hour is 3. Let $Y$ be the number of asthma visits in one hour. Then $Y \sim Poisson(\lambda=3)$

## Exercise 3

What is the probability that in a given hour, exactly 5 asthma patients arrive? Use the `dpois` function to compute your answer and provide a 1-sentence conclusion.

```{r, echo=TRUE}
dpois(x = 5, lambda = 3)
```

## Overview

- Central Limit Theorem 

- Application Exercise

- Why is the Central Limit Theorem important?

## Reading

-   Pagano and Gavreau: Sections 8.1-8.4
-   [OpenIntro Statistics](https://www.openintro.org/go/?id=os4_for_screen_readers&referrer=/book/os/index.php): Section 5.1

## A roadmap ahead

-   In just a few more lectures, we'll have learned enough tools to perform a pretty wide range of analyses and answer associated questions.

-   In the latter half of this course, we'll emphasize specific scientific questions of interest, statistical methods for testing such questions, and translating those results to real-world conclusions.

## A roadmap for today

-   Today's topic is fundamental to making sound statistical inferences and conclusions.

-   The basic idea is called the **central limit theorem**, which states that for **any** distribution with a well-defined mean and variance, the distribution of the means computed from samples of size *n* will be approximately Gaussian (normally distributed).

## Recall: What is statistical inference?

-   Statistical inference is the act of generalizing from a **sample** in order to make conclusions regarding a **population** while quantifying the degree of certainty we have.

-   We are interested in population **parameters**, the "truth," which we do not observe.

-   Instead, we must calculate **statistics** from our sample in order to learn about the parameters.

## The sampling distribution of the mean {.smaller}

Suppose we're interested in the resting heart rate of students in BIOS 600, and are able to do the following:

1.  Take a random sample of size *n* from this population

-   Calculate the mean resting heart rate *in this sample*, $\bar x_1$

::: columns
::: {.column width="50%"}

```{r}
library(tidyverse)
set.seed(123)
#Generate random normal
n <- 20
x <- rnorm(n, mean = 65, sd = 10)

h <- hist(
  x,
  breaks = 30,
  main = "Histogram of Resting Heart Rate (bpm) for Random Sample of 20 people",
  xlab = "Resting Heart Rate (BPM)",
  ylab = "Count"
)
```

:::
::: {.column width="50%"}

```{r}
# Blank scatter plot
plot(
  NA, NA,
  xlim = c(55, 80),
  ylim = c(1, 10),
  xlab = "Average Resting Heart Rate (BPM)",
  ylab = "Count",
  main = "Sampling Distribution of the Mean Resting Heart Rate (BPM)"
)

# Add custom y-axis ticks at 1–5
axis(side = 2, at = 1:10, labels = 1:10)
# Add one dot at (65, 1)
points(65, 1, pch = 19, cex = 1.5)
```

:::
:::


## The sampling distribution of the mean {.smaller}

2.  Put the sample back, take a second random sample of size *n*

-   Calculate the mean resting heart rate from this new sample, $\bar x_2$

3.  Put the sample back, take a third random sample of size *n*

-   Calculate the mean resting heart rate from this sample, too...

... and so on

## The sampling distribution of the mean 

```{r}
# Blank scatter plot
plot(
  NA, NA,
  xlim = c(55, 80),
  ylim = c(1, 10),
  xlab = "Average Resting Heart Rate (BPM)",
  ylab = "Count",
  main = "Sampling Distribution of the Mean Resting Heart Rate (BPM)"
)

# Add custom y-axis ticks at 1–5
axis(side = 2, at = 1:10, labels = 1:10)

# Add one dot at (65, 1)
points(65, 1, pch = 19, cex = 1.5, col="black")
points(60, 1, pch= 19, cex=1.5, col= "blue")
points(62, 1, pch = 19, cex = 1.5, col="hotpink")
```


## The sampling distribution of the mean 

After repeating this many times, we have a dataset that has the sample averages from the population: $\{\bar x_1, \bar x_2, \ldots \}$

-   Can we say anything about the distribution of these sample means?


## The sampling distribution of the mean

```{r}
# --- Population parameters ---
N_pop <- 1000
mu <- 65
sigma <- 10

# Create the population
population <- rnorm(N_pop, mean = mu, sd = sigma)

# --- Sampling parameters ---
n <- 20          # sample size
n_samples <- 300 # number of repeated samples

# Generate sample means
sample_means <- replicate(
  n_samples,
  mean(sample(population, size = n, replace = TRUE))
)

# --- Stack dots into bins ---
breaks <- seq(55, 90, by = 0.5)
bins <- cut(sample_means, breaks = breaks, include.lowest = TRUE)
counts <- table(bins)

# --- Plot ---
plot(
  NA, NA,
  xlim = c(55, 80),
  ylim = c(1, max(counts) + 1),
  xlab = "Sample Mean Resting Heart Rate (BPM)",
  ylab = "Number of Samples",
  main = "Sampling Distribution of the Mean Resting Heart Rate (BPM) - 300 sample means, n = 20"
)

axis(2, at = pretty(0:max(counts)))

# Draw stacked dots
for (i in seq_along(counts)) {
  if (counts[i] > 0) {
    x_center <- mean(breaks[i:(i + 1)])
    points(rep(x_center, counts[i]), 1:counts[i],
           pch = 19, cex = 1.1, col = "steelblue")
  }
}
```
## The central limit theorem 

The **central limit theorem** states that for a population with mean $\mu$ and standard deviation $\sigma$, these three properties hold for the distribution of sample averages $\bar X$:

1.  The mean of the sampling distribution is identical to the population mean $\mu$

2.  The standard deviation of the distribution of the sample averages is $\sigma / \sqrt{n}$, or the **standard error** (SE) of the mean

3.  For *n* large enough (in the limit, as $n \rightarrow \infty$), the shape of the sampling distribution of means is approximately normal

## What is this really saying? 

-   The central limit theorem tells us that **sample averages** are normally distributed, if we have enough data.

-   This is true **even if** our original variables are not normally distributed.

## The sampling distribution of the mean (Smaller n)

```{r}
# --- Population parameters ---
N_pop <- 1000
mu <- 65
sigma <- 10

# Create the population
population <- rnorm(N_pop, mean = mu, sd = sigma)

# --- Sampling parameters ---
n <- 5          # sample size
n_samples <- 300 # number of repeated samples

# Generate sample means
sample_means <- replicate(
  n_samples,
  mean(sample(population, size = n, replace = TRUE))
)

# --- Stack dots into bins ---
breaks <- seq(55, 90, by = 0.5)
bins <- cut(sample_means, breaks = breaks, include.lowest = TRUE)
counts <- table(bins)

# --- Plot ---
plot(
  NA, NA,
  xlim = c(55, 80),
  ylim = c(1, max(counts) + 1),
  xlab = "Sample Mean Resting Heart Rate (BPM)",
  ylab = "Number of Samples",
  main = "Sampling Distribution of the Mean Resting Heart Rate (BPM) - 300 sample means, n = 5"
)

axis(2, at = pretty(0:max(counts)))

# Draw stacked dots
for (i in seq_along(counts)) {
  if (counts[i] > 0) {
    x_center <- mean(breaks[i:(i + 1)])
    points(rep(x_center, counts[i]), 1:counts[i],
           pch = 19, cex = 1.1, col = "steelblue")
  }
}
```

## The sampling distribution of the mean (Larger n)

```{r}
# --- Population parameters ---
N_pop <- 1000
mu <- 65
sigma <- 10

# Create the population
population <- rnorm(N_pop, mean = mu, sd = sigma)

# --- Sampling parameters ---
n <- 100          # sample size
n_samples <- 300 # number of repeated samples

# Generate sample means
sample_means <- replicate(
  n_samples,
  mean(sample(population, size = n, replace = TRUE))
)

# --- Stack dots into bins ---
breaks <- seq(55, 90, by = 0.5)
bins <- cut(sample_means, breaks = breaks, include.lowest = TRUE)
counts <- table(bins)

# --- Plot ---
plot(
  NA, NA,
  xlim = c(55, 80),
  ylim = c(1, max(counts) + 1),
  xlab = "Sample Mean Resting Heart Rate (BPM)",
  ylab = "Number of Samples",
  main = "Sampling Distribution of the Mean Resting Heart Rate (BPM) - 300 sample means, n = 100"
)

axis(2, at = pretty(0:max(counts)))

# Draw stacked dots
for (i in seq_along(counts)) {
  if (counts[i] > 0) {
    x_center <- mean(breaks[i:(i + 1)])
    points(rep(x_center, counts[i]), 1:counts[i],
           pch = 19, cex = 1.1, col = "steelblue")
  }
}
```
## Application Exercise

::: {.callout-tip}
## Application Exercise

For today's participation, go to Canvas and find instructions and template for AE 02.

AE 02 is due on Canvas on no later than **Friday 2/20 at 11:59pm**

:::


## Why the CLT matters {.smaller}

- **Gives us Normality (for free!)**  
  - Even if the population distribution is not normal,  
    the sample mean is approximately normal when sample size is large enough.  

- **Foundation for Inference**  
  - Lets us use normal-based methods for **confidence intervals** and **hypothesis tests**.  
  - Justifies the use of t-tests, z-tests, ANOVAs, regression, etc.  

- **Works Across Many Situations**  
  - Whether data are binary (yes/no), counts, or skewed —  
    sample averages will still tend toward a bell curve.  

- **Practical Takeaway**  
  - We can make powerful statistical conclusions **without needing to know the exact population distribution**.  


## Example: IQ tests

-   IQ tests are designed to have a probability distribution with $\mu$ = 100 and $\sigma$ = 15.

-   Suppose we draw samples of size $n$ = 20 from this population.

-   From the central limit theorem, the distribution of the sample averages will be approximately normal with mean 100 and standard deviation $15/\sqrt{20}$

## IQ tests

-   If the population distribution is normal to begin with, then the distribution of the sample averages will also be exactly normal.

-   If the population distribution is not normal, then the rule of thumb is that we need at least $n = 30$ for the central limit theorem to kick in for approximate normality.

## Example

-   Suppose I give a random sample of $n = 30$ BIOS 600 students an IQ test\*, and the sample average score is 120.

-   Does this mean that BIOS 600 students are smarter than average?

-   I know there are many problems with IQ and IQ testing...bear with me here!

## Example {.smaller}

-   The central limit theorem tells us that the distribution of means of samples of size 30 from this population is also normal, with mean $\mu = 100$ and $SE = \sigma/ \sqrt{n} = 15 \sqrt{30} \approx 2.7$.

-   $Z = \frac{\bar{X} - \mu}{SE}$ is a standard normal random variable, and here $Z \approx 7.3$.

-   The probability of a z-score greater than this is extremely small.

-   What does this mean? (p-values coming soon...)

-   What are the upper and lower limits that enclose 95% of the means for samples of size $n$ drawn from the population? (Confidence intervals coming soon...)


## Recap {.smaller}

-   The central limit theorem is the foundation of statistical inference.

-   The Central Limit Theorem tells us that when you take the average of a large enough sample, that average will follow a normal distribution, even if the original data don’t.

-   This is important because many statistical methods, like confidence intervals and hypothesis tests, rely on normal distributions.

-   Thanks to the CLT, we can apply these methods to real-world data, even if the data itself isn't normally distributed, as long as the sample size is large enough.

## Next class

- Review for Exam 01