---
title: "Lecture 15: Power and sample size"
subtitle: "BIOS 600 - Spring 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Announcements

- No Lab due this week 

- HW 6 can turn in early, firm deadline for **Tuesday, March 24**

- AE 04 due **Tuesday, March 24**

- Professor Howard is out of the office and **will not** have office hours

## Overview

- Why calculate sample size and power?

- How to calculate sample size and power in R

- Application Exercise

## Very optional reading

-   P&G Sections 10.4 and 10.6

-   OI: No corresponding section

## Why calculate sample size and power?

-   To show that under certain required conditions, a hypothesis test has a **good chance of showing the anticipated difference**, if it really exists

-   To be more confident that a null result is not simply a sample of excessive variability

-   To show a funding agency that the study has a reasonable chance of reaching a useful result (leading them to **fund your grant**!)

-   To show that necessary resources (human, animal, financial, time, etc.) will be minimized

## Power {.smaller}

-   Showing adequate statistical power is usually necessary in order to **get funding for research**!

-   **Power** is the probability of rejecting the null hypothesis when it is false (i.e., of avoiding a Type II error)

::: poll
$$\text{Power} = P(\text{reject } H_0 | H_0 \text{ is false})$$
:::

and can be also thought of as the likelihood a planned study will detect a deviation from the null hypothesis if one really exists. Power is a function of:

-   Sample size $n$

-   Deviation from the null one hopes to detect

-   Standard deviation $σ$

-   $\alpha$, the Type I error rate

## Power and Error Rates

- Recall: When we conduct a hypothesis test, there are **four possible outcomes** depending on the true state of the null hypothesis and the decision we make.  

| **Decision**                | **Truth: ($H_0$) True** | **Truth: ($H_0$) False** |
|-----------------------------|--------------------------|---------------------------|
| **Reject ($H_0$)**          | Type I Error ($\alpha$)         | Correct Decision → Power ($1−\beta$) |
| **Fail to Reject ($H_0$)**  | Correct Decision         | Type II Error ($\beta$)          |

## n, detectable difference, variance, and $\alpha$

How do these four considerations affect power?

-   When we design a study, it is not enough to know we have a small probability of rejecting $H_0$ when it is in fact true (i.e. setting $\alpha = 0.05$).

-   We want to know we have a large probability of rejecting the null when it is false. Practically speaking, **power less than 80% is typically considered insufficient to warrant a study**.

## Power

- Recall: In our medical diagnostics lecture, we saw that there is a tradeoff between Type 1 and Type 2 error. Lowering $\alpha$ makes it harder to reject the null hypothesis, which usually reduces power—we’re less likely to catch real effects (true positives).

- Increasing power by tolerating more Type I errors is **not acceptable**. Therefore, we can increase power by

-   Considering **larger deviations from the null** (need to think about clinical/practical importance)

-   Increasing $n$ (good, though not always affordable)

## Interactive visualization

<http://rpsychologist.com/d3/NHST/>

Let's check out this interactive visualization to see how this all works for a one-sample z-test.


## Power Depends On... {.smaller}

| Factor | Change | Effect on Power | Intuition |
|:-------|:-------:|:-------------|:-----------------|
| Sample size ($n$) | $\uparrow$ | $\uparrow$ **Increases** | More data → smaller SE → easier to detect true effects |
| Variability ($\sigma^2$) | $\uparrow$ | $\downarrow$ **Decreases** | More noise → harder to detect signal |
| Effect size ($\delta$) | $\uparrow$ | $\uparrow$ **Increases** | Larger true difference → easier to detect |
| Significance level ($\alpha$) | $\uparrow$ | $\uparrow$ **Increases** | More lenient threshold → higher chance to reject $H_0$ |


## Understanding power

- Holding $n$, variance, and $\alpha$ constant, we have **more power** to detect a **larger difference** compared to a smaller one.

- Holding variance, effect size, and $\alpha$ constant, **increasing sample size ($n$)** increases power — with more data, we’re more likely to detect true effects.

## Understanding Power - Your turn!

::: callout-tip

## Circle One

- Holding $n$, effect size, and $\alpha$ constant, **(higher/lower)** variance decreases power because the signal is harder to detect within the noise.

- Holding $n$, variance, and effect size constant, **(increasing/decreasing)** $\alpha$ increases power — we’re more willing to reject the null, so true effects are easier to detect.

:::


## What is this deviation? {.smaller}

When we calculate power, we need to know the minimum difference from the null mean $\mu_0$ that we wish to detect. We may set up our hypotheses as

$$H_0: \mu = \mu_0$$

$$H_1: \mu \neq \mu_0$$

but need to specify a **minimum detectable difference**, often called $\delta=μ_1−μ_0$ such that we reject $H_0$ with a certain power (usually 80% or 90%) when in fact $μ=μ_1$. We will need a bigger sample size when $\delta$ is small, and fewer subjects when $\delta$ is larger.


## Sample size for for two-sided one-sample test of mean {.smaller}

-   Power and sample size are interrelated.

-   Ideally, when we plan a study, we have a prespecified idea of the minimum difference we want to detect, $δ=μ_1−μ_0$, the standard deviation, $\sigma$, and the power, $1-\beta$, we'd like to have to detect it. In that case, it is straightforward to calculate the required sample size (here given for a one-sample test):


$$n = \left[ \frac{\sigma (z^*_{1-\alpha/2} + z^*_{1-\beta})}{\mu_1-\mu_0}\right]^2$$


-   This can be derived from looking at the rejection regions when $H_0$ and $H_1$ are true, though the details are outside the scope of this class.
-   You can easily calculate power and sample size in R.


## Case study: ultra low dose contraception {.smaller}

-   Suppose you want to ensure a new manufacturer of birth control pills provides the correct dosage of 0.02 $\mu$g estrogen.

-   How many pills do you need to sample in a shipment in order to ensure 80% power to detect a difference of 10% at $\alpha$=0.05, assuming $\sigma$=0.008? (10% of 0.02 is 0.002)

$$n = \left[\frac{\sigma(z^*_{1-\alpha/2} + z^*_{1-\beta})}{\mu_1-\mu_0}\right]^2$$

What are $z^*_{1-\alpha/2}$ and $z^*_{1-\beta}$? These are the $1-\alpha$ and $1-\beta$ quantile cutoffs of a Z-distribution.

## Z cutoffs in R

- $\alpha = 0.05$

- Power = 80%, so Type 2 error rate is 1-.80 = .20 = $\beta$. 

```{r}
#| echo: true
alpha <- 0.05 # type 1 error rate
beta <- .20 # type 2 error rate
qnorm(1-alpha/2); qnorm(1-beta)
```

## Plugging in our values {.smaller}

$= \left[\frac{.008(1.96 + 0.84)}{0.002}\right]^2$

$= 125.44$

Using a normal approximation, we would need 126 pills sampled to detect the desired difference of 10%, given $\sigma = 0.008$, at 80% power.

- To be conservative, always **round up for sample size calculations**.

## We use the t-distribution in practice

- In the example above, we used the z-distribution. However, we use the $t$ distribution in practice instead of the $z$. 

- This is because the population standard deviation ($\sigma$) is **rarely known**, so we estimate it from the sample.  

- The *t*-distribution accounts for this extra uncertainty.

- When sample sizes are **small or moderate**, using the *t*-distribution provides more accurate power estimates, since it better reflects the variability from estimating $\sigma$.

## In R

```{r}
#| echo: true
power.t.test(delta = 0.002,
             sd = 0.008,
             sig.level = 0.05,
             power = 0.8,
             type = "one.sample")
```

Using the t-distribution, we need 128 pills sampled to detect the desired difference.


## Example: Power Calculation

- Researchers are testing whether a new low-sodium diet reduces **systolic blood pressure (SBP)** compared to a control diet.  

- Previous studies suggest an average reduction of **8 mmHg** with a standard deviation of **10 mmHg**.  

- They plan to recruit **20 participants per group** and use a **two-sided t-test** at the 0.05 level.

**Question:** What statistical *power* will they have to detect this difference?

## In R

\footnotesize

```{r}
#| echo: true
# Power calculation for two-sample t-test
power.t.test(n = 20, 
             delta = 8, 
             sd = 10, 
             sig.level = 0.05, 
             type = "two.sample", 
             alternative = "two.sided")
```
## Interpretation

**Interpretation**: With 20 participants per group, the study has about 69% power to detect an 8 mmHg difference.

- In other words, if the true effect exists, there’s roughly a 7-in-10 chance we’ll correctly reject the null hypothesis.

## Example 2: Sample size calculation

- Suppose we want to test whether a new medication lowers LDL cholesterol compared to a placebo. 

- From pilot data, we expect a difference of 15 mg/dL with a standard deviation of 20 mg/dL. 

- We want 90% power at a 0.05 significance level for a two-sided test. 

**Question**: How many participants per group are needed?

## In R

\footnotesize

```{r}
#| echo: true
power.t.test(power = 0.90,
             delta = 15,
             sd = 20,
             sig.level = 0.05,
             type = "two.sample",
             alternative = "two.sided")
```

## Interpretation

39 individuals per group are needed (rounding up) to achieve 90% power to detect a 15 mg/dL difference in LDL cholesterol between treatments. 

## Application Exercise

::: callout-tip

## AE 04

Head to Canvas and begin working on Application Exercise (AE) 04: Power and Sample Size Calculation. 

- AE 04 is due **Tuesday 3/24 at 11:59pm**. 

- (Turn in a PDF on Canvas.)

:::

## Next Class

- Comparing two continuous variables

- Enjoy Spring Break!!