---
title: "Lecture 13: Introduction to linear regression"
subtitle: "BIOS 600 - Summer Session II 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Announcements

- Lab 3 due **tonight** by 11:59 PM

- HW 13 due **tomorrow** by 11:59 PM

## Revisiting lab from Friday

- In lab Friday we talked about two additional concepts
  - Partial correlation
  - Spearman's correlation coefficients

## Pearson vs. Spearman

- A linear relationship is always monotonic, but monotonic doesn't have to be linear (e.g., exponential growth)
- Because it uses ranks, Spearman is less sensitive to outliers than Pearson
- Rule of thumb: 
  - Data is linear → use Pearson (more power)
  - Data is skewed, has outliers, or the relationship is curved but consistently increasing/decreasing → use Spearman

```{r}
#| echo: false
#| eval: true

library(ggplot2)
library(tidyverse)
library(tidymodels)
library(kableExtra)
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 60)   # outlier at end
y <- c(2, 4, 5, 4, 5, 7, 8, 9, 10, 11)
```

## Example: Pearson vs. Spearman

```{r}
#| echo: false
#| eval: true
#| fig-alt: "A scatterplot with 10 points; nine of them form a positive-sloped line with low x-values; one has a much higher x value and is an outlier."
tibble(x, y) |>
  ggplot(aes(x, y)) +
  geom_point(size = 4) +
  labs(x = "X", y = "Y") +
  theme_minimal()
```

## Example: Pearson vs. Spearman 

Pearson Correlation

```{r}
#| echo: true
cor(x, y, method = "pearson")
```

Spearman Correlation

```{r}
#| echo: true
cor(x, y, method = "spearman")
```


## Introduction to linear regression

-   Now let's talk about other methods for exploring the relationships between variables

## Reading

-   P&G Chapter 18

-   OI: Chapter 8

## Back to Lab 1...

```{r}
library(tidyverse)
cdc <- read.csv("https://aggreenbean.github.io/Bios600/labs/data/cdc_cleaned.csv")
```

```{r}
#| fig-alt: "A scatterplot with x-axis labeled Adequate aerobic activity (%) and y-axis labeled Obesity (%). The points have a general downward trend."
cdc %>%
  ggplot(aes(x = Obesity, y = Exercise)) +
  geom_point() +
  theme_bw() +
  labs(x = "Adequate aerobic activity (%)", 
       y = "Obesity (%)")
```

## Correlation and regression {.smaller}

-   Correlation measures the direction and strength of linear relationships

-   Regression can be used to further quantify these relationships

-   Often used to test whether there is an association between two variables

-   A **regression** line summarizes the relationship between explanatory/predictor variables (e.g., Exercise % or $X$) and response/outcome variables (e.g., Obesity % or $Y$)

-   The fitted regression line can be used to **predict the outcome** for a given set of predictor values. Our predictions do have error, called **residuals** (vertical distance between observation and the regression line).

-   The **least-squares regression line** minimizes the sum of the squared errors.

## Regression {.smaller}

-   In regression, we use one variable (or more) to try to predict values of another.

-   In simple linear regression, one variable $Y$ is the **response** or **outcome** or **dependent variable** and the other $X$ is the **predictor** or **explanatory** variable or **independent** variable.

-   **Important:** The regression of $Y$ on $X$ is **not equal** to the regression of $X$ on $Y$

-   The regression of $Y$ on $X$ can be used to predict $Y$ based on fixed values of $X$.

## Back to Lab 1... {.smaller}

```{r}
#| fig-alt: "A scatterplot with x-axis labeled Adequate aerobic activity (%) and y-axis labeled Obesity (%). The points have a general downward trend. A negatively-sloped blue line is superimposed over the scatterplot to emphasize this trend."
cdc %>%
  ggplot(aes(x = Obesity, y = Exercise)) +
  geom_point() +
  theme_bw() +
  labs(x = "Adequate aerobic activity (%)", 
       y = "Obesity (%)") +
  geom_smooth(method = "lm", se = F) 
```

\footnotesize

-   We see the regression line relating exercise and obesity %

-   Points represent actual data values; note that the line is not a perfect fit

## Code for previous slide

```{r}
#| echo: true
#| eval: false
cdc |>
  ggplot(aes(x = Obesity, y = Exercise)) +
  geom_point() +
  theme_bw() +
  labs(x = "Adequate aerobic activity (%)", 
       y = "Obesity (%)") +
  geom_smooth(method = "lm", se = F) 
```

## The linear model

The model is given by $y_i=β_0+β_1x_i+ϵ_i$, where

-   $y_i$ is the outcome (dependent variable) of interest

-   $\beta_0$ is the intercept parameter

-   $\beta_1$ is the slope parameter

-   $x_i$ is the predictor variable

-   $\epsilon_i$ is the error (like $\beta_0$ and $\beta_1$, it is **not observed**)

The $i$ is an index of observations. So, each observation of the outcome is related to an observation of that individual's value of the predictor.

## The least squares model {.smaller}

Fitted line equation:

$$\hat y_i = \hat \beta_0 +\hat \beta_1 x_i$$

Residual:

$$\hat \epsilon_i = y_i - (\hat \beta_0 +\hat \beta_1 x_i) = y_i - \hat y_i$$

Least squares selects $\hat β_0$ and $\hat β_1$ such that the **error sum of squares**

$$\sum_{i=1}^n \hat \epsilon_i^2 = \sum_{i=1}^n (y_i - \hat y_i)^2$$

is minimized.

## Review slope-intercept form of a line {.smaller}

Recall the slope-intercept form of a line, $y=mx+b$.

-   $b$ is the y-intercept, or where the line crosses the y-axis

    -   It is the predicted value of $y$ when $x = 0$

-   $m$ is the slope

    -   Which tells us the **predicted increase** in $y$ when $x$ **changes by 1 unit**

## Review: formula of a line

::: callout-tip

## Review

Consider the line $y=3x+1$. 

- What value does $y$ take when $x=0$?

- As $x$ increases by 1 unit, how much do we expect $y$ to increase by?
:::

-   In statistics, we often represent the **slope** with $β_1$ and the **intercept** with $β_0$, and these values are usually not known but must be estimated.

## Model output {.smaller}

```{r echo = T}
fit <- lm(Obesity ~ Exercise, data = cdc)
summary(fit)
```

## A neater way: tidymodels

- **tidymodels:** "tidyverse for modeling" — R packages that brings a consistent, unified workflow to modeling
- Once you learn this workflow, it scales to more complex models down the road without learning a whole new system
- Gives us the `tidy()` function, which converts messy `lm()` output into a clean, organized data frame
- Much easier to read and work with than the default `summary()` output

## A neater way: kableExtra

- **kableExtra** is an R package for turning raw R output into clean, readable tables in your document
- Today we're using it to format regression results so they're easier to read and interpret
- Small amount of code, much more professional-looking and easier to read output

## A neater way

```{r}
#| echo: true
library(tidymodels)
library(kableExtra)

tidy(fit) |>
  kable()
```

- Can we do better?


## Even better

```{r}
#| echo: true
tidy(fit) |>
  kable(digits = 3) # round to 3 digits
```

- Note: p-values are not actually 0. They are just quite small so they get rounded to zero in the table. 

::: callout-tip
- If you include a regression model in your labs and homework, display the output with `tidy()` and `kable()` like this!
:::

## Intercept and slope

-   The **intercept** is the estimated mean outcome when the predictor is zero (does this make sense in context?)

-   The **slope** is the expected change in the outcome corresponding to a **one-unit change** in the predictor


**Slope ($\beta_1$) Interpretation**: For each percentage increase in `Exercise`, we expect the `Obesity` percentage to decrease by .48 on average

::: callout-tip
For all interpretations, remember to **include units**!
:::

## Example

For two states, identical except one has Exercise % = 45.4 and the other has Exercise % = 46.4

- State A: $\hat{y}_{A} = 54.79 + (-0.49 \times 45.4) = 32.54$
- State B: $\hat{y}_{B} = 54.79 + (-0.49 \times 46.4) = 32.05$
- Difference: $32.54 - 32.05 = 0.49$

Increasing Exercise % by one unit changes predicted Obesity % by exactly $\hat{\beta}_1 = -0.49$

## Intercept interpretation

- **The technical intercept interpretation**: When zero percent of the region gets adequate `Exercise`, we would expect the percent of the region that is obese to be 54.79%. 

- Does it make sense to interpret the intercept here?

- Only interpret the intercept in a regression model if:

  - the predictor can **feasibly** take values equal to or near zero, or
  - there are values near zero in the observed data
  
## Let's check

```{r}
#| echo: true
#| fig-alt: "A histogram titled Histogram of exercise, with x axis labeled Exercise and y-axis labeled frequency. There are six bins, each covering five percentage points from 35 to 65. The distribution appears normal."
hist(cdc$Exercise)
```

## Let's check

```{r}
#| echo: true
min(cdc$Exercise)
```

- Since the predictor (`Exercise`) does not feasibly take on values equal to zero (and we also checked that the lowest value in the dataset is 37.4%), we should not interpret the intercept. 

## Model assumptions {.smaller}

Assumptions of the model $y_i = \beta_0 + \beta_1 x_1 + \epsilon_i$ include:

-   The outcomes $y_i$ are independent. This is violated if our study contains repeated outcome measures on an individual, if siblings are enrolled, etc.

-   A linear relationship holds (though we can relax this to some extent)

-   For a specified value of $x$, which is measured without error, $y_i∼N(β_0+β_1x_i,σ^2)$. Note that this implies $ϵ_i∼N(0,σ^2)$.

-   The variance $σ^2$ is constant across all values of $x$; this is called **homogeneity of variance**

## Hypothesis testing {.smaller}

-   The primary hypothesis test of interest is usually $H_0:β_1=0$ (There is no association between exercise and obesity percentage) versus $H_1:β_1≠0$ (there is an association between exercise and obesity percentage.)

-   Under the null hypothesis, we can use a t-test to test this
      - Same logic, same mechanics as the t-tests we've already seen — testing how likely is it see an estimate of $\beta_1$ this extreme or more extreme under the $H_{0}: \beta_1=0$

- R automatically tests this with the `lm()` function, and provides the t-statistic and p-value in the output.

```{r}
tidy(fit) |>
  kable(digits = 3) # round to 3 digits
```

::: {.callout-tip appearance="simple"}
Given our t-statistic of -7.7 and p-value of \<0.001, what might we conclude in our data?
:::

## Some additional output {.smaller}

```{r}
#| echo: true
summary(fit)$r.squared
```

- The $R-squared$ value indicates that exercise explains 55% of the variance of obesity. 

- $R^2$ tells us about how much of the variability in obesity is explained by exercise; as it gets close to 1, the points get tighter around the regression line.

- $R^2$ ranges from 0 to 1

## Predicted values 

- The regression equation can be used to get predicted means at any value of $x$. 

- $\hat y_i=\hat β_0+ \hat β_1x_i$, where we call $\hat y$ the predicted value of $y$ at a given value of $x$ (what happened to the error term?). 

## Predicted values {.smaller}

- For the CDC data that predicts a state's obesity percentage by its adequate exercise percentage, we have

$$\hat y_i = \hat β_0 - \hat β_1 x_i = 54.79 - 0.49 x_i$$

So, the predicted mean obesity percentage for a state where 50% of residents get adequate exercise is

$$\hat y_i = 54.79 - 0.49 (50) $$

We would predict an obesity percentage of approximately 30%. Note the importance of getting the units correct.

## Return to the assumptions of linear regression {.smaller}

::: callout-tip

## Review 

1.  Independent observations

2.  Linear relationship between predictor and response

3.  Conditional on predictor(s), the response follows a normal distribution (normally distributed residuals)

4.  Equal variances (homogeneity of variance in the residuals)

:::

- How can we check the assumptions to ensure they are met? 

- We can do so by "**running regression diagnostics**" - which means checking the assumptions within our regression model. 

## Regression diagnostics: independence {.smaller}

We just have to think through this one, unfortunately.

- The **independence assumption** means that each observation provides *unique information* — no observation should influence another.

- In other words, the residuals (errors) should not be correlated with each other.

::: callout-tip

## Common violations of independence
- Repeated measurements on the same subject over time  
  (e.g., daily blood pressure readings for the same patient)
  
- Clustered or grouped data  
  (e.g., patients within the same hospital, students within a classroom)
  
- Time-series data  
  (e.g., monthly pollution levels, daily step counts)

:::

## Regression diagnostics: equal variance; linearity

-   To check equal variances and linearity, we can use a plot of the residuals $\hat ϵ_i$ by the predicted (or fitted) values $\hat y_i$.

-   We expect to see evenly-spaced dots along the y axis (equal variances), with symmetrically distributed observations around the y axis (linearity). Patterns or trends are evidence something is wrong. (Drawing on next slide.)

## Regression diagnostics: equal variances

```{r}
#| echo: false
#| eval: true
#| fig-alt: "A scatterplot with x-axis labeled Fitted and y-axis labeled Residuals. There is a dashed red line at y=0, and the x-axis ranges from 1 to 10. The points of the scatterplot are closer to the dashed line for lower x values and spread out further for higher x-values."
library(ggplot2)
set.seed(42)

n <- 100
fitted_vals <- seq(1, 10, length.out = n)

# --- Plot 1: Non-equal variance (funnel shape) ---
residuals_hetero <- rnorm(n, mean = 0, sd = 0.3 * fitted_vals)

df_hetero <- data.frame(fitted = fitted_vals, residuals = residuals_hetero)

df_hetero |> ggplot(aes(x = fitted, y = residuals)) +
  geom_point(shape = 16, alpha = 0.6, size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
  geom_vline(xintercept = mean(fitted_vals), color = "darkred", linewidth = 1) +
  labs(x = expression(hat(y)[i] ~ "(Fitted)"),
    y = expression(hat(epsilon)[i] ~ "(Residuals)")
  ) +
  theme_minimal()
```
## Regression diagnostics: equal variances
```{r}
#| fig-alt: "A scatterplot with x-axis labeled Fitted and y-axis labeled Residuals. There is a dashed red line at y=0, and the x-axis ranges from 1 to 10. The points of the scatterplot form a parabola."
# --- Plot 2: Violates linearity (curved/quadratic pattern) ---
residuals_nonlinear <- -1.5 * (fitted_vals - 5.5)^2 + 10 + rnorm(n, sd = 5)
residuals_nonlinear <- residuals_nonlinear - mean(residuals_nonlinear)

df_nonlinear <- data.frame(fitted = fitted_vals, residuals = residuals_nonlinear)

df_nonlinear |> ggplot(aes(x = fitted, y = residuals)) +
  geom_point(shape = 16, alpha = 0.6, size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
  labs(
    x = expression(hat(y)[i] ~ "(Fitted)"),
    y = expression(hat(epsilon)[i] ~ "(Residuals)")
  ) +
  theme_minimal()
```

## Regression diagnostic: equal variances

```{r}
#| echo: false
#| eval: true
#| fig-alt: "A scatterplot with x-axis labeled Fitted and y-axis labeled Residuals. There is a dashed red line at y=0, and the x-axis ranges from 1 to 10. The points of the scatterplot are equally split above and below the dashed line for all x values."
# --- Plot 3: Equal variance (looks good!) ---
residuals_homo <- rnorm(n, mean = 0, sd = 1.5)

df_homo <- data.frame(fitted = fitted_vals, residuals = residuals_homo)

df_homo |> ggplot(aes(x = fitted, y = residuals)) +
  geom_point(shape = 16, alpha = 0.6, size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
  labs(
    x = expression(hat(y)[i] ~ "(Fitted)"),
    y = expression(hat(epsilon)[i] ~ "(Residuals)")
  ) +
  theme_minimal()
```
## Regression diagnostics: normality of errors

-   Use a histogram of the residuals to check normality. There are more advanced methods and plots, but we won't talk about them here

## Checking assumptions for our model...

```{r}
#| echo: false
#| warning: false
#| message: false
cdc <- read.csv("https://karamccor.github.io/b6002/labs/data/cdc_cleaned.csv")
```

```{r echo = T}
#| fig-alt: "A scatterplot with x-axis labeled fitted and y-axis labeled residuals, with a dashed black line at y = 0. The points are generally equally above and below the dashed line."
fit <- lm(Obesity ~ Exercise, data = cdc)
plot(fit$fitted.values, fit$residuals)
abline(h = 0,lty = 2)
```

## Checking assumptions for our model...

```{r}
#| echo: true
#| fig-alt: "A histogram with x-axis labeled Residuals and y-axis labeled Frequency. The distribution appears normal."
# check normality of residuals
hist(fit$residuals)
```

## Recap

- What is a linear regression model? How do we fit & visualize the model in R?

- Slope, intercept and interpretations

- Understand when it's appropriate to interpret the intercept

- Interpret $R^2$ from regression output

- How to check regression diagnostics

## Next class

- Multiple regression
