---
title: "LTA analysis - Descriptive Stats"
author: "Annie Green Howard"
date: today
format: pdf
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(writexl)
library(slca)
library(LCPA)
library(openxlsx)
options(scipen = 999)

todd_dir <- "Z:\\LTA project\\Data\\"
```

## Reading in dataset from 2009

2009 dataset now includes a total of 8,702 observations (matching analytic dataset in published paper) and includes a total of 79 variables - including indexno

```{r}
lca_pcp_2009 <- read.table(
  file.path(todd_dir, "lca_pcp_2009_1.24.23.dat"),
  sep = ",",
  header = FALSE,
  fill = TRUE,
  comment.char = "",
  quote = ""
)

#Changing all values of -9999 to missing
lca_pcp_2009[lca_pcp_2009==-9999] <- NA
```


```{r}
#| include: false
#| warning: false
#| message: false
vars_2009 <- c("indexno","white", "black", "american_native", "asian_or_pi", "hispanicrace", "other_race", "age",
          "obstetrics", "patient_care_hours", "prenatal", "group_fam", "group_int", "group_obgyn",
          "group_ped", "group_ger", "ruca_urban", "ruca_lrural", "ruca_srural", "hpsa_not", "hpsa_full",
          "hpsa_part", "adi_state", "setting_amb", "setting_hosp", "setting_medsch", "setting_nurs",
          "setting_other", "setting_solo", "setting_missing", "female", "gen_x", "gen_y", "gen_boom_lead",
          "gen_boom_trail", "gen_ww2", "race_unknown", "school_img", "school_in_state", "school_out_hbcu",
          "school_out_nonhbcu", "zadi_state", "zhours", "practice_yrs", "_0to5", "_6to10", "_11to20",
          "_21to30", "_31more", "fwhite", "fblack", "fan", "fasian", "fhispanic", "fother", "funknown",
          "mwhite", "mblack", "man", "masian", "mhispanic", "mother", "munknown", "fgen_y", "fgen_x", "fgen_boom_lead",
          "fgen_boom_trail", "fgen_ww2", "mgen_y", "mgen_x", "mgen_boom_lead", "mgen_boom_trail",
          "mgen_ww2", "age35below", "age36to45", "age46to55", "age56to65", "age66over", "group")

names(lca_pcp_2009) <- vars_2009
```

## Reading in dataset from 2019 

2019 dataset now includes a total of 7,929 observations (matching analytic dataset in published paper) and includes a total of 79 variables - including indexno

```{r}
lca_pcp_2019 <- read.table(
  file.path(todd_dir, "lca_pcp2019_1.24.23.dat"),
  sep = ",",
  header = FALSE,
  fill = TRUE,
  comment.char = "",
  quote = ""
)

#Changing all values of -9999 to missing
lca_pcp_2019[lca_pcp_2019==-9999] <- NA

names(lca_pcp_2019) <- vars_2009
```

## Reformatting variables for descriptive tables

Now we need to format the variables to be categorical variables in the 2009 dataset

```{r}
lca_pcp_2009 <- lca_pcp_2009 %>%
  mutate(specialty = case_when(
    group_fam   == 1 ~ "Family Medicine",
    group_int   == 1 ~ "General Internal Medicine",
    group_obgyn == 1 ~ "General Obstetrics and Gynecology",
    group_ped   == 1 ~ "General Pediatrics",
    group_ger   == 1 ~ "Geriatrics",
    TRUE             ~ "Missing"
  ),
  employment_setting = case_when(
    setting_amb   == 1 ~ "Ambulatory Care",
    setting_hosp  == 1 ~ "Hospital",
    setting_medsch == 1 ~ "Medical School/Parent University",
    setting_nurs  == 1 ~ "Nursing Home/Extended Care",
    setting_other == 1 ~ "Other", 
    setting_solo == 1 ~ "Solo practioner",
    setting_missing == 1 ~ "Missing",
    TRUE             ~ "Missing Value"
  ),
  urban_rural = case_when(
    ruca_urban  == 1 ~ "Urban",
    ruca_lrural == 1 ~ "Large Rural",
    ruca_srural == 1 ~ "Small Rural",
    TRUE             ~ "Missing"
  ),
  race_ethnicity = case_when(
    white == 1           ~ "Non-Hispanic White",
    black == 1           ~ "Non-Hispanic Black",
    american_native == 1 ~ "American Indian/Alaskan Native",
    asian_or_pi == 1     ~ "Asian/Pacific Islander",
    hispanicrace == 1    ~ "Hispanic",
    other_race == 1      ~ "Other",
    race_unknown == 1    ~ "Unknown",
    TRUE                 ~ "Missing"
  ),
  generational = case_when(
    gen_x          == 1 ~ "Gen X (1965-1980)",
    gen_y          == 1 ~ "Gen Y (1981-2000)",
    gen_boom_lead  == 1 ~ "Baby Boomer, Leading Edge (1946-1954)",
    gen_boom_trail == 1 ~ "Baby Boomer, Trailing Edge (1955-1964)",
    gen_ww2        == 1 ~ "WWII (1928-1945)",
    TRUE                ~ "Missing"
  ))
```

Now we need to format the variables to be categorical variables 

```{r}
lca_pcp_2019 <- lca_pcp_2019 %>%
  mutate(specialty = case_when(
    group_fam   == 1 ~ "Family Medicine",
    group_int   == 1 ~ "General Internal Medicine",
    group_obgyn == 1 ~ "General Obstetrics and Gynecology",
    group_ped   == 1 ~ "General Pediatrics",
    group_ger   == 1 ~ "Geriatrics",
    TRUE             ~ "Missing"
  ),
  employment_setting = case_when(
    setting_amb   == 1 ~ "Ambulatory Care",
    setting_hosp  == 1 ~ "Hospital",
    setting_medsch == 1 ~ "Medical School/Parent University",
    setting_nurs  == 1 ~ "Nursing Home/Extended Care",
    setting_other == 1 ~ "Other", 
    setting_solo == 1 ~ "Solo practioner",
    setting_missing == 1 ~ "Missing",
    TRUE             ~ "Missing Value"
  ),
  urban_rural = case_when(
    ruca_urban  == 1 ~ "Urban",
    ruca_lrural == 1 ~ "Large Rural",
    ruca_srural == 1 ~ "Small Rural",
    TRUE             ~ "Missing"
  ),
  race_ethnicity = case_when(
    white == 1           ~ "Non-Hispanic White",
    black == 1           ~ "Non-Hispanic Black",
    american_native == 1 ~ "American Indian/Alaskan Native",
    asian_or_pi == 1     ~ "Asian/Pacific Islander",
    hispanicrace == 1    ~ "Hispanic",
    other_race == 1      ~ "Other",
    race_unknown == 1    ~ "Unknown",
    TRUE                 ~ "Missing"
  ),
  generational = case_when(
    gen_x          == 1 ~ "Gen X (1965-1980)",
    gen_y          == 1 ~ "Gen Y (1981-2000)",
    gen_boom_lead  == 1 ~ "Baby Boomer, Leading Edge (1946-1954)",
    gen_boom_trail == 1 ~ "Baby Boomer, Trailing Edge (1955-1964)",
    gen_ww2        == 1 ~ "WWII (1928-1945)",
    TRUE                ~ "Missing"
  ))
```

## Creating merged dataset 

First we want to rename all variables, except indexno - our ID - with subscript denoting year. Also adding an indicator variable to denote whether the indexno is in 2009 and/or in 2019. 

```{r}
lca_pcp_2009_renamed <- lca_pcp_2009 %>%
  rename_with(~ paste0(.x, "_2009"), -indexno, -in_2009) %>%
  mutate(in_2009 = 1L) 

lca_pcp_2019_renamed <- lca_pcp_2019 %>%
  rename_with(~ paste0(.x, "_2019"), -indexno, -in_2019) %>%
  mutate(in_2019 = 1L) 
```

Now merging dataset - including all indexno in **either** 2009 and/or 2019

```{r}
lca_pcp_merged <- lca_pcp_2009_renamed %>%
  full_join(lca_pcp_2019_renamed, by = "indexno") %>%
  mutate(
    in_2009 = coalesce(in_2009, 0L),
    in_2019 = coalesce(in_2019, 0L)
  )
```

## Now looking at descriptive statistics

```{r}
table(lca_pcp_merged$in_2009, lca_pcp_merged$in_2019)
```

```{r}
summary_table <- lca_pcp_merged %>%
  select(indexno, in_2009, in_2019, all_of(num_vars)) %>%
  pivot_longer(
    cols = all_of(num_vars),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    year = case_when(
      str_ends(variable, "_2009") ~ "2009",
      str_ends(variable, "_2019") ~ "2019"
    ),
    base_var = str_remove(variable, "_2009|_2019")
  ) %>%
  group_by(base_var) %>%
  mutate(
    is_binary = all(na.omit(value) %in% c(0, 1))
  ) %>%
  ungroup() %>%
  mutate(
    group = case_when(
      in_2009 == 1 & in_2019 == 0 & year == "2009" ~ "2009_only",
      in_2009 == 1 & in_2019 == 1 & year == "2009" ~ "both_2009",
      in_2009 == 1 & in_2019 == 1 & year == "2019" ~ "both_2019",
      in_2009 == 0 & in_2019 == 1 & year == "2019" ~ "2019_only",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(group)) %>%
  group_by(base_var, is_binary, group) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    sd   = sd(value, na.rm = TRUE),
    n1   = sum(value == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    value_display = if_else(
      is_binary,
      paste0(
        round(mean * 100), "% (N = ", n1, ")"
      ),
      paste0(
        round(mean, 1),
        " (",
        round(sd, 1),
        ")"
      )
    )
  ) %>%
  select(base_var, group, value_display) %>%
  pivot_wider(
    names_from = group,
    values_from = value_display
  ) %>%
  arrange(base_var)

summary_table
```

Formatting this table so this can be output - including column sample sizes

```{r}
n_2009_only <- sum(lca_pcp_merged$in_2009 == 1 & lca_pcp_merged$in_2019 == 0)
n_both      <- sum(lca_pcp_merged$in_2009 == 1 & lca_pcp_merged$in_2019 == 1)
n_2019_only <- sum(lca_pcp_merged$in_2009 == 0 & lca_pcp_merged$in_2019 == 1)

summary_table_n <- summary_table

colnames(summary_table_n) <- c(
  "Variable",
  paste0("2009 only (N = ", n_2009_only, ")"),
  paste0("Both years – 2009 (N = ", n_both, ")"),
  paste0("Both years – 2019 (N = ", n_both, ")"),
  paste0("2019 only (N = ", n_2019_only, ")")
)
```

## Output summary table 

```{r}
write.csv(
  summary_table_n,
  file = "lca_pcp_summary_table_041423.csv",
  row.names = FALSE
)
```

