---
title: "LTA analysis - Descriptive Stats"
author: "Annie Green Howard"
date: today
format: pdf
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(writexl)
library(slca)
library(LCPA)
library(openxlsx)
library(ggplot2)
library(ggalluvial)

options(scipen = 999)

todd_dir <- "Z:\\LTA project\\Data\\"
```

## Reading in dataset from 2009

2009 dataset now includes a total of 8,702 observations (matching analytic dataset in published paper) and includes a total of 79 variables - including indexno

```{r}
lca_pcp_2009 <- read.table(
  file.path(todd_dir, "lca_pcp_2009_1.24.23.dat"),
  sep = ",",
  header = FALSE,
  fill = TRUE,
  comment.char = "",
  quote = ""
)

#Changing all values of -9999 to missing
lca_pcp_2009[lca_pcp_2009==-9999] <- NA
```


```{r}
#| include: false
#| warning: false
#| message: false
vars_2009 <- c("indexno","white", "black", "american_native", "asian_or_pi", "hispanicrace", "other_race", "age",
          "obstetrics", "patient_care_hours", "prenatal", "group_fam", "group_int", "group_obgyn",
          "group_ped", "group_ger", "ruca_urban", "ruca_lrural", "ruca_srural", "hpsa_not", "hpsa_full",
          "hpsa_part", "adi_state", "setting_amb", "setting_hosp", "setting_medsch", "setting_nurs",
          "setting_other", "setting_solo", "setting_missing", "female", "gen_x", "gen_y", "gen_boom_lead",
          "gen_boom_trail", "gen_ww2", "race_unknown", "school_img", "school_in_state", "school_out_hbcu",
          "school_out_nonhbcu", "zadi_state", "zhours", "practice_yrs", "_0to5", "_6to10", "_11to20",
          "_21to30", "_31more", "fwhite", "fblack", "fan", "fasian", "fhispanic", "fother", "funknown",
          "mwhite", "mblack", "man", "masian", "mhispanic", "mother", "munknown", "fgen_y", "fgen_x", "fgen_boom_lead",
          "fgen_boom_trail", "fgen_ww2", "mgen_y", "mgen_x", "mgen_boom_lead", "mgen_boom_trail",
          "mgen_ww2", "age35below", "age36to45", "age46to55", "age56to65", "age66over", "group")

names(lca_pcp_2009) <- vars_2009
```

## Reading in dataset from 2019 

2019 dataset now includes a total of 7,929 observations (matching analytic dataset in published paper) and includes a total of 79 variables - including indexno

```{r}
lca_pcp_2019 <- read.table(
  file.path(todd_dir, "lca_pcp2019_1.24.23.dat"),
  sep = ",",
  header = FALSE,
  fill = TRUE,
  comment.char = "",
  quote = ""
)

#Changing all values of -9999 to missing
lca_pcp_2019[lca_pcp_2019==-9999] <- NA

names(lca_pcp_2019) <- vars_2009
```

## Creating merged dataset 

First we want to rename all variables, except indexno - our ID - with subscript denoting year. Also adding an indicator variable to denote whether the indexno is in 2009 and/or in 2019. 

```{r}
lca_pcp_2009_renamed <- lca_pcp_2009 %>%
  rename_with(~ paste0(.x, "_2009"), -indexno, -in_2009) %>%
  mutate(in_2009 = 1L) 

lca_pcp_2019_renamed <- lca_pcp_2019 %>%
  rename_with(~ paste0(.x, "_2019"), -indexno, -in_2019) %>%
  mutate(in_2019 = 1L) 
```

Now merging dataset - including all indexno in **either** 2009 and/or 2019

```{r}
lca_pcp_merged <- lca_pcp_2009_renamed %>%
  full_join(lca_pcp_2019_renamed, by = "indexno") %>%
  mutate(
    in_2009 = coalesce(in_2009, 0L),
    in_2019 = coalesce(in_2019, 0L)
  )
```

## Now looking at descriptive statistics

```{r}
table(lca_pcp_merged$in_2009, lca_pcp_merged$in_2019)
```

```{r}
num_vars <- lca_pcp_merged %>%
  select(where(is.numeric)) %>%
  select(-indexno, -in_2009, -in_2019) %>%
  names()

summary_table <- lca_pcp_merged %>%
  select(indexno, in_2009, in_2019, all_of(num_vars)) %>%
  pivot_longer(
    cols = all_of(num_vars),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    year = case_when(
      str_ends(variable, "_2009") ~ "2009",
      str_ends(variable, "_2019") ~ "2019"
    ),
    base_var = str_remove(variable, "_2009|_2019")
  ) %>%
  group_by(base_var) %>%
  mutate(
    is_binary = all(na.omit(value) %in% c(0, 1))
  ) %>%
  ungroup() %>%
  mutate(
    group = case_when(
      in_2009 == 1 & in_2019 == 0 & year == "2009" ~ "2009_only",
      in_2009 == 1 & in_2019 == 1 & year == "2009" ~ "both_2009",
      in_2009 == 1 & in_2019 == 1 & year == "2019" ~ "both_2019",
      in_2009 == 0 & in_2019 == 1 & year == "2019" ~ "2019_only",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(group)) %>%
  group_by(base_var, is_binary, group) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    sd   = sd(value, na.rm = TRUE),
    n1   = sum(value == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    value_display = if_else(
      is_binary,
      paste0(
        round(mean * 100), "% (N = ", n1, ")"
      ),
      paste0(
        round(mean, 1),
        " (",
        round(sd, 1),
        ")"
      )
    )
  ) %>%
  select(base_var, group, value_display) %>%
  pivot_wider(
    names_from = group,
    values_from = value_display
  ) %>%
  arrange(base_var)

summary_table
```

Formatting this table so this can be output - including column sample sizes

```{r}
n_2009_only <- sum(lca_pcp_merged$in_2009 == 1 & lca_pcp_merged$in_2019 == 0)
n_both      <- sum(lca_pcp_merged$in_2009 == 1 & lca_pcp_merged$in_2019 == 1)
n_2019_only <- sum(lca_pcp_merged$in_2009 == 0 & lca_pcp_merged$in_2019 == 1)

summary_table_n <- summary_table

colnames(summary_table_n) <- c(
  "Variable",
  paste0("2009 only (N = ", n_2009_only, ")"),
  paste0("Both years – 2009 (N = ", n_both, ")"),
  paste0("Both years – 2019 (N = ", n_both, ")"),
  paste0("2019 only (N = ", n_2019_only, ")")
)
```

## Output summary table 

```{r}
write.csv(
  summary_table_n,
  file = "lca_pcp_summary_table_041623.csv",
  row.names = FALSE
)
```

## Comparisons of specialty

We first want to define categories - first for speciality

```{r}
lca_pcp_merged <- lca_pcp_merged %>%
  mutate(
    specialty_2009 = case_when(
      group_fam_2009   == 1 ~ "Family Medicine",
      group_int_2009   == 1 ~ "General Internal Medicine",
      group_obgyn_2009 == 1 ~ "General Obstetrics and Gynecology",
      group_ped_2009   == 1 ~ "General Pediatrics",
      group_ger_2009   == 1 ~ "Geriatrics",
      TRUE                  ~ "Missing"
    ),
    specialty_2019 = case_when(
      group_fam_2019   == 1 ~ "Family Medicine",
      group_int_2019   == 1 ~ "General Internal Medicine",
      group_obgyn_2019 == 1 ~ "General Obstetrics and Gynecology",
      group_ped_2019   == 1 ~ "General Pediatrics",
      group_ger_2019   == 1 ~ "Geriatrics",
      TRUE                  ~ "Missing"
    )
  )
```

Now I want to look at table, showing movement allowing for row and column totals, ideally looking at those only in 2009 and 2019

```{r}
specialty_levels <- c(
  "Family Medicine",
  "General Internal Medicine",
  "General Obstetrics and Gynecology",
  "General Pediatrics",
  "Geriatrics",
  "Missing"
)


lca_pcp_merged_combine <- lca_pcp_merged %>%
  filter(in_2009 == 1 & in_2019 == 1) %>%
  mutate(
    specialty_2009 = factor(specialty_2009, levels = specialty_levels),
    specialty_2019 = factor(specialty_2019, levels = specialty_levels)
  )


tab <- table(
  lca_pcp_merged_combine$specialty_2009,
  lca_pcp_merged_combine$specialty_2019
)

tab_with_margins <- addmargins(tab)


tab_df <- as.data.frame.matrix(tab_with_margins) %>%
  tibble::rownames_to_column(var = "2009 Specialty")

# Rename the total column
names(tab_df)[ncol(tab_df)] <- "2019 Total"

# Rename the total row
tab_df$`2009 Specialty` <- ifelse(
  tab_df$`2009 Specialty` == "Sum",
  "2009 Total",
  tab_df$`2009 Specialty`
)


write.csv(
  tab_df,
  file = "specialty_transition_table_2009_2019.csv",
  row.names = FALSE
)

```

Now creating the sanskey plot

```{r}
sankey_data <- lca_pcp_merged %>%
  filter(in_2009 == 1, in_2019 == 1) %>%
  count(
    specialty_2009,
    specialty_2019,
    name = "n"
  )

p_sankey_special <- suppressWarnings(
  ggplot(
    sankey_data,
    aes(
      axis1 = specialty_2009,
      axis2 = specialty_2019,
      y = n
    )
  ) +
    geom_alluvium(aes(fill = specialty_2009), width = 1/12, alpha = 0.8) +
    geom_stratum(width = 1/12, fill = "grey90", color = "black") +
    geom_text(
      stat = "stratum",
      aes(label = stringr::str_wrap(after_stat(stratum), width = 20)),
      size = 3
    ) +
    scale_x_discrete(
      limits = c("2009 Specialty", "2019 Specialty"),
      expand = c(0.15, 0.05)
    ) +
    labs(
      title = "Physician Specialty Transitions from 2009 to 2019",
      subtitle = "Restricted to physicians observed in both years",
      y = "Number of Physicians",
      x = NULL,
      fill = "2009 Specialty"
    ) +
    theme_minimal() +
    theme(
      legend.position = "right",
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank()
    )
)



ggsave(
  filename = "specialty_sankey_2009_2019.png",
  plot = p_sankey_special,
  width = 9,
  height = 6,
  dpi = 300,
  bg = "white"
)
```


Then for setting

```{r}
lca_pcp_merged <- lca_pcp_merged %>%
  mutate(
    employment_setting_2009 = case_when(
      setting_amb_2009     == 1 ~ "Ambulatory Care",
      setting_hosp_2009    == 1 ~ "Hospital",
      setting_medsch_2009  == 1 ~ "Medical School",
      setting_nurs_2009    == 1 ~ "Nursing Home",
      setting_other_2009   == 1 ~ "Other",
      setting_solo_2009    == 1 ~ "Solo Practitioner",
      setting_missing_2009 == 1 ~ "Missing",
      TRUE                      ~ "Missing Value"
    ),
    employment_setting_2019 = case_when(
      setting_amb_2019     == 1 ~ "Ambulatory Care",
      setting_hosp_2019    == 1 ~ "Hospital",
      setting_medsch_2019  == 1 ~ "Medical School",
      setting_nurs_2019    == 1 ~ "Nursing Home",
      setting_other_2019   == 1 ~ "Other",
      setting_solo_2019    == 1 ~ "Solo Practitioner",
      setting_missing_2019 == 1 ~ "Missing",
      TRUE                      ~ "Missing Value"
    )
  )
```

Creating descriptive tables

```{r}
employment_setting_levels <- c(
  "Ambulatory Care",
  "Hospital",
  "Medical School",
  "Nursing Home",
  "Solo Practitioner",
  "Other",
  "Missing",
  "Missing Value"
)


lca_pcp_merged_combine <- lca_pcp_merged %>%
  filter(in_2009 == 1 & in_2019 == 1) %>%
  mutate(
    employment_setting_2009 =
      factor(employment_setting_2009, levels = employment_setting_levels),
    employment_setting_2019 =
      factor(employment_setting_2019, levels = employment_setting_levels)
  )


tab_setting <- table(
  lca_pcp_merged_combine$employment_setting_2009,
  lca_pcp_merged_combine$employment_setting_2019
)

tab_setting_with_margins <- addmargins(tab_setting)


tab_setting_df <- as.data.frame.matrix(tab_setting_with_margins) %>%
  tibble::rownames_to_column(var = "2009 Employment Setting")

# Rename the total column
names(tab_setting_df)[ncol(tab_setting_df)] <- "2019 Total"

# Rename the total row
tab_setting_df$`2009 Employment Setting` <- ifelse(
  tab_setting_df$`2009 Employment Setting` == "Sum",
  "2009 Total",
  tab_setting_df$`2009 Employment Setting`
)


write.csv(
  tab_setting_df,
  file = "employment_setting_transition_table_2009_2019.csv",
  row.names = FALSE
)


```


```{r}
sankey_setting <- lca_pcp_merged %>%
  filter(in_2009 == 1, in_2019 == 1) %>%
  count(
    employment_setting_2009,
    employment_setting_2019,
    name = "n"
  )

p_setting <- suppressWarnings(
  ggplot(
    sankey_setting,
    aes(
      axis1 = employment_setting_2009,
      axis2 = employment_setting_2019,
      y     = n
    )
  ) +
    geom_alluvium(
      aes(fill = employment_setting_2009),
      width = 1/12,
      alpha = 0.8
    ) +
    geom_stratum(
      width = 1/12,
      fill  = "grey90",
      color = "black"
    ) +

    geom_text(
      stat = "stratum",
      aes(label = stringr::str_wrap(after_stat(stratum), 22)),
      size = 3.2,
      lineheight = 0.9,   # ✅ THIS is the single‑spacing fix
      fill = "white",
      label.size = 0.2,
      alpha = 0.9
    ) +
    scale_x_discrete(
      limits = c("2009 Setting", "2019 Setting"),
      expand = c(0.15, 0.25)
    ) +
    coord_cartesian(clip = "off") +
    labs(
      title    = "Employment Setting Transitions from 2009 to 2019",
      subtitle = "Restricted to physicians observed in both years",
      y        = "Number of Physicians",
      x        = NULL,
      fill     = "2009 Employment Setting"
    ) +
    theme_minimal() +
    theme(
      legend.position      = "right",
      panel.grid.major.x  = element_blank(),
      panel.grid.minor    = element_blank(),
      plot.margin         = margin(t = 10, r = 80, b = 10, l = 10)
    )
)


ggsave(
  filename = "employment_setting_sankey_2009_2019.png",
  plot     = p_setting,
  width    = 10,
  height   = 6,
  dpi      = 300,
  bg       = "white"
)

```

Now for urban rural

```{r}
lca_pcp_merged <- lca_pcp_merged %>%
  mutate(
    urban_rural_2009 = case_when(
      ruca_urban_2009  == 1 ~ "Urban",
      ruca_lrural_2009 == 1 ~ "Large Rural",
      ruca_srural_2009 == 1 ~ "Small Rural",
      TRUE                  ~ "Missing"
    ),
    urban_rural_2019 = case_when(
      ruca_urban_2019  == 1 ~ "Urban",
      ruca_lrural_2019 == 1 ~ "Large Rural",
      ruca_srural_2019 == 1 ~ "Small Rural",
      TRUE                  ~ "Missing"
    )
  )
```

Add descriptive table

```{r}
# --- Define full set of levels (including Missing) ---
urban_rural_levels <- c(
  "Urban",
  "Large Rural",
  "Small Rural",
  "Missing"
)

# --- Restrict to those observed in both years and lock factor levels ---
lca_pcp_merged_combine <- lca_pcp_merged %>%
  filter(in_2009 == 1 & in_2019 == 1) %>%
  mutate(
    urban_rural_2009 = factor(urban_rural_2009, levels = urban_rural_levels),
    urban_rural_2019 = factor(urban_rural_2019, levels = urban_rural_levels)
  )

# --- Create transition table ---
tab_urban <- table(
  lca_pcp_merged_combine$urban_rural_2009,
  lca_pcp_merged_combine$urban_rural_2019
)

# --- Add row and column totals ---
tab_urban_with_margins <- addmargins(tab_urban)

# --- Convert to data frame and label totals ---
tab_urban_df <- as.data.frame.matrix(tab_urban_with_margins) %>%
  rownames_to_column(var = "2009 Urban–Rural")

# Rename total column
names(tab_urban_df)[ncol(tab_urban_df)] <- "2019 Total"

# Rename total row
tab_urban_df$`2009 Urban–Rural` <- ifelse(
  tab_urban_df$`2009 Urban–Rural` == "Sum",
  "2009 Total",
  tab_urban_df$`2009 Urban–Rural`
)

# --- Write to CSV ---
write.csv(
  tab_urban_df,
  file = "urban_rural_transition_table_2009_2019.csv",
  row.names = FALSE
)
```


Adding Sanskey plot 

```{r}

sankey_urban <- lca_pcp_merged %>%
  filter(in_2009 == 1, in_2019 == 1) %>%
  count(
    urban_rural_2009,
    urban_rural_2019,
    name = "n"
  )

p_urban <- suppressWarnings(
  ggplot(
    sankey_urban,
    aes(
      axis1 = urban_rural_2009,
      axis2 = urban_rural_2019,
      y     = n
    )
  ) +
    geom_alluvium(
      aes(fill = urban_rural_2009),
      width = 1/12,
      alpha = 0.8
    ) +
    geom_stratum(
      width = 1/12,
      fill  = "grey90",
      color = "black"
    ) +

    # ---- Stratum labels (no boxes) ----
    geom_text(
      stat = "stratum",
      aes(label = str_wrap(after_stat(stratum), 18)),
      size = 3.2,
      lineheight = 0.9,
      color = "black"
    ) +

    scale_x_discrete(
      limits = c("2009 Urban–Rural", "2019 Urban–Rural"),
      expand = c(0.15, 0.25)
    ) +
    coord_cartesian(clip = "off") +
    labs(
      title    = "Urban–Rural Transitions from 2009 to 2019",
      subtitle = "Restricted to physicians observed in both years",
      y        = "Number of Physicians",
      x        = NULL,
      fill     = "2009 Urban–Rural"
    ) +
    theme_minimal() +
    theme(
      legend.position     = "right",
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      plot.margin        = margin(t = 10, r = 80, b = 10, l = 80)
    )
)


ggsave(
  filename = "urban_rural_sankey_2009_2019.png",
  plot     = p_urban,
  width    = 9,
  height   = 6,
  dpi      = 300,
  bg       = "white"
)
```

## Doing this table for race/ehtniciy

Then for setting





race_unknown


```{r}
lca_pcp_merged <- lca_pcp_merged %>%
  mutate(

    ## ---- Race (2009) ----
    race_2009 = case_when(
      white_2009            == 1 ~ "Non-Hispanic White",
      black_2009            == 1 ~ "Non-Hispanic Black",
      american_native_2009  == 1 ~ "American Indian/Alaskan Native",
      asian_or_pi_2009      == 1 ~ "Asian/Pacific Islander",
      hispanicrace_2009     == 1 ~ "Hispanic",
      other_race_2009       == 1 ~ "Other",
      race_unknown_2009     == 1 ~ "Unknown",
      TRUE                       ~ "Missing Value"
    ),

    ## ---- Race (2019) ----
    race_2019 = case_when(
      white_2019            == 1 ~ "Non-Hispanic White",
      black_2019            == 1 ~ "Non-Hispanic Black",
      american_native_2019  == 1 ~ "American Indian/Alaskan Native",
      asian_or_pi_2019      == 1 ~ "Asian/Pacific Islander",
      hispanicrace_2019     == 1 ~ "Hispanic",
      other_race_2019       == 1 ~ "Other",
      race_unknown_2019     == 1 ~ "Unknown",
      TRUE                       ~ "Missing Value"
    )

  )
```

Now doing tabble and figure

```{r}
library(dplyr)
library(tidyr)
library(tibble)

# --- Define full set of race levels (must match how you defined race_2009/2019) ---
race_levels <- c(
  "Non-Hispanic White",
  "Non-Hispanic Black",
  "American Indian/Alaskan Native",
  "Asian/Pacific Islander",
  "Hispanic",
  "Other",
  "Unknown",
  "Missing Value"
)

# --- Restrict to those observed in both years and lock factor levels ---
lca_pcp_merged_combine <- lca_pcp_merged %>%
  filter(in_2009 == 1 & in_2019 == 1) %>%
  mutate(
    race_2009 = factor(race_2009, levels = race_levels),
    race_2019 = factor(race_2019, levels = race_levels)
  )

# --- Create transition table ---
tab_race <- table(
  lca_pcp_merged_combine$race_2009,
  lca_pcp_merged_combine$race_2019
)

# --- Add row and column totals ---
tab_race_with_margins <- addmargins(tab_race)

# --- Convert to data frame and label totals ---
tab_race_df <- as.data.frame.matrix(tab_race_with_margins) %>%
  rownames_to_column(var = "2009 Race")

# Rename total column
names(tab_race_df)[ncol(tab_race_df)] <- "2019 Total"

# Rename total row
tab_race_df$`2009 Race` <- ifelse(
  tab_race_df$`2009 Race` == "Sum",
  "2009 Total",
  tab_race_df$`2009 Race`
)

# --- Write to CSV ---
write.csv(
  tab_race_df,
  file = "race_transition_table_2009_2019.csv",
  row.names = FALSE
)
```

