---
title: "LTA analysis"
author: "Annie Green Howard"
format: html
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(writexl)
library(slca)
library(LCPA)
library(purrr)
library(tidyverse)
library(writexl)
library(slca)
library(LCPA)
library(openxlsx)
library(ggplot2)
library(ggalluvial)
library(purrr)
options(scipen = 999)

todd_dir <- "Z:\\LTA project\\Data\\"
```

## Reading in dataset from 2009

2009 dataset now includes a total of 8,702 observations (matching analytic dataset in published paper) and includes a total of 79 variables - including indexno

```{r}
lca_pcp_2009 <- read.table(
  file.path(todd_dir, "lca_pcp_2009_1.24.23.dat"),
  sep = ",",
  header = FALSE,
  fill = TRUE,
  comment.char = "",
  quote = ""
)

#Changing all values of -9999 to missing
lca_pcp_2009[lca_pcp_2009==-9999] <- NA
```


```{r}
#| include: false
#| warning: false
#| message: false
vars_2009 <- c("indexno","white", "black", "american_native", "asian_or_pi", "hispanicrace", "other_race", "age",
          "obstetrics", "patient_care_hours", "prenatal", "group_fam", "group_int", "group_obgyn",
          "group_ped", "group_ger", "ruca_urban", "ruca_lrural", "ruca_srural", "hpsa_not", "hpsa_full",
          "hpsa_part", "adi_state", "setting_amb", "setting_hosp", "setting_medsch", "setting_nurs",
          "setting_other", "setting_solo", "setting_missing", "female", "gen_x", "gen_y", "gen_boom_lead",
          "gen_boom_trail", "gen_ww2", "race_unknown", "school_img", "school_in_state", "school_out_hbcu",
          "school_out_nonhbcu", "zadi_state", "zhours", "practice_yrs", "_0to5", "_6to10", "_11to20",
          "_21to30", "_31more", "fwhite", "fblack", "fan", "fasian", "fhispanic", "fother", "funknown",
          "mwhite", "mblack", "man", "masian", "mhispanic", "mother", "munknown", "fgen_y", "fgen_x", "fgen_boom_lead",
          "fgen_boom_trail", "fgen_ww2", "mgen_y", "mgen_x", "mgen_boom_lead", "mgen_boom_trail",
          "mgen_ww2", "age35below", "age36to45", "age46to55", "age56to65", "age66over", "group")

names(lca_pcp_2009) <- vars_2009
```

## Reading in dataset from 2019 

2019 dataset now includes a total of 7,929 observations (matching analytic dataset in published paper) and includes a total of 79 variables - including indexno

```{r}
lca_pcp_2019 <- read.table(
  file.path(todd_dir, "lca_pcp2019_1.24.23.dat"),
  sep = ",",
  header = FALSE,
  fill = TRUE,
  comment.char = "",
  quote = ""
)

#Changing all values of -9999 to missing
lca_pcp_2019[lca_pcp_2019==-9999] <- NA

names(lca_pcp_2019) <- vars_2009
```

## Creating merged dataset 

First we want to rename all variables, except indexno - our ID - with subscript denoting year. Also adding an indicator variable to denote whether the indexno is in 2009 and/or in 2019. 

```{r}
lca_pcp_2009_renamed <- lca_pcp_2009 %>%
  rename_with(~ paste0(.x, "_2009"), -indexno, -in_2009) %>%
  mutate(in_2009 = 1L) 

lca_pcp_2019_renamed <- lca_pcp_2019 %>%
  rename_with(~ paste0(.x, "_2019"), -indexno, -in_2019) %>%
  mutate(in_2019 = 1L) 
```

Now merging dataset - including all indexno in **either** 2009 and/or 2019

```{r}
lca_pcp_merged <- lca_pcp_2009_renamed %>%
  full_join(lca_pcp_2019_renamed, by = "indexno") %>%
  mutate(
    in_2009 = coalesce(in_2009, 0L),
    in_2019 = coalesce(in_2019, 0L)
  )

#Keeping only those in 2009 and 2019
lca_pcp_merged_combine <- lca_pcp_merged %>%
  filter(in_2009 == 1 & in_2019 == 1) 
```

## Running with 500 random starts

```{r}
fits <- map(1:500, function(s) {
  set.seed(s)

  slca(
    l_09[4] ~ group_fam_2009 + group_int_2009 + group_obgyn_2009 +
      group_ped_2009 + group_ger_2009 +
      setting_amb_2009 + setting_hosp_2009 + setting_medsch_2009 +
      setting_nurs_2009 + setting_solo_2009 + setting_other_2009 +
      obstetrics_2009 + prenatal_2009 +
      ruca_urban_2009 + ruca_lrural_2009 + ruca_srural_2009,

    l_19[4] ~ group_fam_2019 + group_int_2019 + group_obgyn_2019 +
      group_ped_2019 + group_ger_2019 +
      setting_amb_2019 + setting_hosp_2019 + setting_medsch_2019 +
      setting_nurs_2019 + setting_solo_2019 + setting_other_2019 +
      obstetrics_2019 + prenatal_2019 +
      ruca_urban_2019 + ruca_lrural_2019 + ruca_srural_2019,

    l_09 ~ l_19
  ) %>%
    estimate(
      data = lca_pcp_merged_combine,
      start = "random",
      control = list(em.total = 1e-6)
    )
})
```

Now checking model fit

```{r}
logliks <- sapply(fits, logLik)

#Get log likilhood values
ll_tab       <- sort(table(logliks), decreasing = TRUE)
head(ll_tab, 20)

#Checking if these are sorted
ll_tab_round <- sort(table(round(logliks, 6)), decreasing = TRUE)
head(ll_tab_round, 20)
```

```{r}
# tabulate exact values
ll_tab <- table(logliks)

# find the modal log-likelihood *as character*
modal_char <- names(ll_tab)[which.max(ll_tab)]
modal_char

# select indices by character comparison
best_indices <- which(as.character(logliks) == modal_char)
best_indices

best_model <- fits[[ best_indices[1] ]]
```

```{r}
logLik(best_model)
param(best_model)
```

## Save param(best_model)

```{r}
# Extract parameter estimates
par_df <- as.data.frame(param(best_model))

# Write to CSV
write.csv(
  par_df,
  "param_best_model.csv",
  row.names = FALSE
)

```


## Getting dataset with class assignments

```{r}
# Get probabilities for each class per individual
post_probs <- predict(best_model, type = "posterior")

post_marginal <- post_probs$marginal

# Getting 2009 classes
post_09 <- post_marginal[[1]]
# Getting 2019 classes
post_19 <- post_marginal[[2]]

## Assiging IDs to these individuals
class_09_ml <- apply(post_09, 1, which.max)
class_19_ml <- apply(post_19, 1, which.max)

# Merging with origina dataaset
lca_with_classes <- cbind(
  lca_pcp_merged_combine,
  class_09_ml = class_09_ml,
  class_19_ml = class_19_ml
)

write.csv(
  lca_with_classes,
  "data_set_with_classes.csv",
  row.names = FALSE
)

```

