---
title: "LTA analysis"
author: "Annie Green Howard"
format: html
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(writexl)
library(slca)
library(LCPA)
library(purrr)
```

## Looking at the 2009 dataset and getting it to be similar to 2019 dataset

The 2009 dataset does include closer to the true number, including 8,699 individuals. Same sample size for both 1.24.23 version and the v2 version. These datasets both include indexno variable. We cannot seem to get to exactly 8,702 individuals, in Table 1 in published paper, was there exclusion criteria we care about there where we might also want to include them in the descriptive table for LTA work?

```{r}
#Reading in the version 2 datasets for 2009 but this seems to be an older version, dated 11/22/2023
lca_pcp_2009v2      <- read.table("Original Data/lca_pcp_2009v2.dat", header = TRUE, sep = ",", skip = 2) 
#Changing all values of -9999 to missing
lca_pcp_2009v2[lca_pcp_2009v2==-9999] <- NA

#Reading in the datasets for 2009 and 2019 as of 1.24.23
lca_pcp_2009_12423  <- read.table("Original Data/lca_pcp_2009_1.24.23.dat", header = TRUE, sep = ",", skip = 2) 
#Changing all values of -9999 to missing
lca_pcp_2009_12423[lca_pcp_2009_12423==-9999] <- NA

vars_2009 <- c("indexno","white", "black", "american_native", "asian_or_pi", "hispanicrace", "other_race", "age",
          "obstetrics", "patient_care_hours", "prenatal", "group_fam", "group_int", "group_obgyn",
          "group_ped", "group_ger", "ruca_urban", "ruca_lrural", "ruca_srural", "hpsa_not", "hpsa_full",
          "hpsa_part", "adi_state", "setting_amb", "setting_hosp", "setting_medsch", "setting_nurs",
          "setting_other", "setting_solo", "setting_missing", "female", "gen_x", "gen_y", "gen_boom_lead",
          "gen_boom_trail", "gen_ww2", "race_unknown", "school_img", "school_in_state", "school_out_hbcu",
          "school_out_nonhbcu", "zadi_state", "zhours", "practice_yrs", "_0to5", "_6to10", "_11to20",
          "_21to30", "_31more", "fwhite", "fblack", "fan", "fasian", "fhispanic", "fother", "funknown",
          "mwhite", "mblack", "man", "masian", "mhispanic", "mother", "munknown", "fgen_y", "fgen_x", "fgen_boom_lead",
          "fgen_boom_trail", "fgen_ww2", "mgen_y", "mgen_x", "mgen_boom_lead", "mgen_boom_trail",
          "mgen_ww2", "age35below", "age36to45", "age46to55", "age56to65", "age66over", "group")

names(lca_pcp_2009_12423) <- vars_2009
```

Now we need to format the variables to be categorical variables 

```{r}
lca_pcp_2009_12423 <- lca_pcp_2009_12423 %>%
  mutate(specialty = case_when(
    group_fam   == 1 ~ "Family Medicine",
    group_int   == 1 ~ "General Internal Medicine",
    group_obgyn == 1 ~ "General Obstetrics and Gynecology",
    group_ped   == 1 ~ "General Pediatrics",
    group_ger   == 1 ~ "Geriatrics",
    TRUE             ~ "Missing"
  ),
  employment_setting = case_when(
    setting_amb   == 1 ~ "Ambulatory Care",
    setting_hosp  == 1 ~ "Hospital",
    setting_medsch == 1 ~ "Medical School/Parent University",
    setting_nurs  == 1 ~ "Nursing Home/Extended Care",
    setting_other == 1 ~ "Other", 
    setting_solo == 1 ~ "Solo practioner",
    setting_missing == 1 ~ "Missing",
    TRUE             ~ "Missing Value"
  ),
  urban_rural = case_when(
    ruca_urban  == 1 ~ "Urban",
    ruca_lrural == 1 ~ "Large Rural",
    ruca_srural == 1 ~ "Small Rural",
    TRUE             ~ "Missing"
  ),
  race_ethnicity = case_when(
    white == 1           ~ "Non-Hispanic White",
    black == 1           ~ "Non-Hispanic Black",
    american_native == 1 ~ "American Indian/Alaskan Native",
    asian_or_pi == 1     ~ "Asian/Pacific Islander",
    hispanicrace == 1    ~ "Hispanic",
    other_race == 1      ~ "Other",
    race_unknown == 1    ~ "Unknown",
    TRUE                 ~ "Missing"
  ),
  generational = case_when(
    gen_x          == 1 ~ "Gen X (1965-1980)",
    gen_y          == 1 ~ "Gen Y (1981-2000)",
    gen_boom_lead  == 1 ~ "Baby Boomer, Leading Edge (1946-1954)",
    gen_boom_trail == 1 ~ "Baby Boomer, Trailing Edge (1955-1964)",
    gen_ww2        == 1 ~ "WWII (1928-1945)",
    TRUE                ~ "Missing"
  ))
```

## Looking at the 2019 dataset

Note that the variables in the 2009 dataset are the exact same variables in the 2019 dataset. 

There are a total of 7,926 individuals in the 2019 dataset. 

```{r}
lca_pcp2019_12423   <- read.table("Original Data/lca_pcp2019_1.24.23.dat", header = TRUE, sep = ",", skip = 2) 
#Changing all values of -9999 to missing
lca_pcp2019_12423[lca_pcp2019_12423==-9999] <- NA

names(lca_pcp2019_12423) <- vars_2009
```

Now we need to format the variables to be categorical variables 

```{r}
lca_pcp2019_12423 <- lca_pcp2019_12423 %>%
  mutate(specialty = case_when(
    group_fam   == 1 ~ "Family Medicine",
    group_int   == 1 ~ "General Internal Medicine",
    group_obgyn == 1 ~ "General Obstetrics and Gynecology",
    group_ped   == 1 ~ "General Pediatrics",
    group_ger   == 1 ~ "Geriatrics",
    TRUE             ~ "Missing"
  ),
  employment_setting = case_when(
    setting_amb   == 1 ~ "Ambulatory Care",
    setting_hosp  == 1 ~ "Hospital",
    setting_medsch == 1 ~ "Medical School/Parent University",
    setting_nurs  == 1 ~ "Nursing Home/Extended Care",
    setting_other == 1 ~ "Other", 
    setting_solo == 1 ~ "Solo practioner",
    setting_missing == 1 ~ "Missing",
    TRUE             ~ "Missing Value"
  ),
  urban_rural = case_when(
    ruca_urban  == 1 ~ "Urban",
    ruca_lrural == 1 ~ "Large Rural",
    ruca_srural == 1 ~ "Small Rural",
    TRUE             ~ "Missing"
  ),
  race_ethnicity = case_when(
    white == 1           ~ "Non-Hispanic White",
    black == 1           ~ "Non-Hispanic Black",
    american_native == 1 ~ "American Indian/Alaskan Native",
    asian_or_pi == 1     ~ "Asian/Pacific Islander",
    hispanicrace == 1    ~ "Hispanic",
    other_race == 1      ~ "Other",
    race_unknown == 1    ~ "Unknown",
    TRUE                 ~ "Missing"
  ),
  generational = case_when(
    gen_x          == 1 ~ "Gen X (1965-1980)",
    gen_y          == 1 ~ "Gen Y (1981-2000)",
    gen_boom_lead  == 1 ~ "Baby Boomer, Leading Edge (1946-1954)",
    gen_boom_trail == 1 ~ "Baby Boomer, Trailing Edge (1955-1964)",
    gen_ww2        == 1 ~ "WWII (1928-1945)",
    TRUE                ~ "Missing"
  ))
```

## Creating combined datasets

Changing variable names to _2009 or _2019 depending on what dataset they are from 

```{r}
colnames(lca_pcp_2009_12423) <- ifelse(colnames(lca_pcp_2009_12423) == "indexno", "indexno", paste0(colnames(lca_pcp_2009_12423), "_2009"))
colnames(lca_pcp2019_12423)  <- ifelse(colnames(lca_pcp2019_12423) == "indexno", "indexno", paste0(colnames(lca_pcp2019_12423), "_2019"))
```

Now merging the dataset keeping only individuals in both datasets

```{r}
merged_lta <- merge(lca_pcp_2009_12423, lca_pcp2019_12423, by = "indexno")
```

This results in a total of 4,614 individuals. Could we look at descriptive statistics for combinations of variables across time

In particular we are not sure about specialty, given questions have changed over time so this shifted broad patterns of what specialties show up. Not necessarily a change of practice, but a change in how they reported their practice.

```{r}
table(merged_lta$specialty_2009,merged_lta$specialty_2019)
```
## Now running LCA among these individuals

Note packages such as **slca** and **LCPA** in R might be appropriate in these cases. 

Here we are holding the item response probabilites stable over time

```{r}
model1 <- 
    slca(l_09[4] ~ group_fam_2009 + group_int_2009 + group_obgyn_2009 
     + group_ped_2009 + group_ger_2009 + setting_amb_2009 + setting_hosp_2009 + setting_medsch_2009
     + setting_nurs_2009 + setting_solo_2009 + setting_other_2009 
     + obstetrics_2009 + prenatal_2009 + ruca_urban_2009 + ruca_lrural_2009 + ruca_srural_2009,
    l_19[4] ~ group_fam_2019 + group_int_2019 + group_obgyn_2019 
     + group_ped_2019 + group_ger_2019 + setting_amb_2019 + setting_hosp_2019 + setting_medsch_2019
     + setting_nurs_2019 + setting_solo_2019 + setting_other_2019 
     + obstetrics_2019 + prenatal_2019 + ruca_urban_2019 + ruca_lrural_2019 + ruca_srural_2019, 
    l_09 ~ l_19, constraints = c("l_09", "l_19")) %>%
  estimate(data = merged_lta, control=list(em.total=1e-6))
summary(model1)
param(model1)
```

## Now not holding constnat 


Here we are holding the item response probabilites stable over time

```{r}
model3 <-
  slca(
    l_09[4] ~ 
      group_fam_2009 + group_int_2009 + group_obgyn_2009 + 
      group_ped_2009 + group_ger_2009 + 
      setting_amb_2009 + setting_hosp_2009 + setting_medsch_2009 +
      setting_nurs_2009 + setting_solo_2009 + setting_other_2009 +
      obstetrics_2009 + prenatal_2009 +
      ruca_urban_2009 + ruca_lrural_2009 + ruca_srural_2009,

    l_19[4] ~ 
      group_fam_2019 + group_int_2019 + group_obgyn_2019 +
      group_ped_2019 + group_ger_2019 +
      setting_amb_2019 + setting_hosp_2019 + setting_medsch_2019 +
      setting_nurs_2019 + setting_solo_2019 + setting_other_2019 +
      obstetrics_2019 + prenatal_2019 +
      ruca_urban_2019 + ruca_lrural_2019 + ruca_srural_2019,

    l_09 ~ l_19
  ) %>%
estimate(
  data = merged_lta,
  start = "random",
  control = list(
    em.total = 1e-6,
    em.iter.max = 5000
  )
)
```
## Checking model convergence

```{r}
logLik(model3)
param(model3)
```


## Now looking at just 2009

```{r}
model3_2009 <- 
    slca(l_09[4] ~ group_fam_2009 + group_int_2009 + group_obgyn_2009 
     + group_ped_2009 + group_ger_2009 + setting_amb_2009 + setting_hosp_2009 + setting_medsch_2009
     + setting_nurs_2009 + setting_solo_2009 + setting_other_2009 
     + obstetrics_2009 + prenatal_2009 + ruca_urban_2009 + ruca_lrural_2009 + ruca_srural_2009) %>%
  estimate(data = merged_lta, control=list(em.total=1e-6))
summary(model3_2009)
param(model3_2009)
```

## Now tyring to handle categories

```{r}
library(purrr)

fits <- map(1:500, function(s) {
  set.seed(s)

  slca(
    l_09[4] ~ group_fam_2009 + group_int_2009 + group_obgyn_2009 +
      group_ped_2009 + group_ger_2009 +
      setting_amb_2009 + setting_hosp_2009 + setting_medsch_2009 +
      setting_nurs_2009 + setting_solo_2009 + setting_other_2009 +
      obstetrics_2009 + prenatal_2009 +
      ruca_urban_2009 + ruca_lrural_2009 + ruca_srural_2009,

    l_19[4] ~ group_fam_2019 + group_int_2019 + group_obgyn_2019 +
      group_ped_2019 + group_ger_2019 +
      setting_amb_2019 + setting_hosp_2019 + setting_medsch_2019 +
      setting_nurs_2019 + setting_solo_2019 + setting_other_2019 +
      obstetrics_2019 + prenatal_2019 +
      ruca_urban_2019 + ruca_lrural_2019 + ruca_srural_2019,

    l_09 ~ l_19
  ) %>%
    estimate(
      data = merged_lta,
      start = "random",
      control = list(em.total = 1e-6)
    )
})
```

Now checking model fit

```{r}
logliks <- sapply(fits, logLik)
table(logliks)
```

```{r}
best_model <- fits[ which.max(logliks)]
```

```{r}
logLik(best_model)
param(best_model)
```


