---
title: "Application Exercise 3 - t-Tests"
author: "YOUR NAME HERE"
date: "Today's Date Here"
format: html
---

## Setup

Update your name and the date in the YAML above.

## Learning Goal

Today you will practice using `t.test()` in R. You will analyze data from a school athletics department, which wants to test the effectiveness of a new type of training by comparing the average times of 10 runners in the 100 meters in seconds before and after the new training is implemented.

Our dataset in R is called `training`, which can be loaded with the following code chunk. Here, `echo = F` means that this chunk will be evaluated when you render your Quarto document, but it won't be visible in the final document.

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

library(tidyverse)
subject <- 1:30
before <- c(12.9, 13.5, 12.8, 15.6, 17.2, 19.2, 12.6, 15.3, 14.4, 11.3,
            11.6, 12.8, 11, 20.4, 15.2, 20.1, 14.5, 14, 13.1, 10.9,
            12.3, 11.8, 15, 13.9, 17.5, 16.1, 14.3, 13.5, 14.9, 11.8)
after <- c(12, 12.2, 11.2, 13.0, 15, 15.8, 12.2, 13.4, 12.9, 11,
           12.1, 12, 11.5, 18.3, 14.5, 17.1, 12.9, 13.5, 12.8, 10.9,
           12.5, 11.4, 15.3, 13.2, 15.9, 15.4, 12.5, 13.1, 13.1, 11.5)
training <- data.frame(subject, before, after)

```

## Question 1

To run a paired t-test of the before and after times, you use the same function as for a two-sample t-test, with the additional parameter `paired = TRUE`.

```{r}

## complete the function as appropriate to run a paired t-test between 
## the before and after times

t.test(training$before,
       training$after, 
       mu = ,
       alternative = ,
       paired = ,
       var.equal = ,
       conf.level = 
       )

```

What conclusion can you make from this t-test?

## Question 2

### a. 

Now, create a new variable in `training` called `diff`, which is the difference between `before` and `after`.

```{r}

## Complete the code to create the new variable

training <- training |>
  

```

### b. 

Run a one-sample t-test on `training$diff`, where the null hypothesis is $H_0$: $\mu_{diff} = 0$.

```{r}



```

What conclusion can you make from this t-test?

### c. 

Compare your results from the test in Question 1 to your results from the test in Question 2b. What do you observe?

