---
title: "Exam Review"
date: "September 18, 2025"
format: html
---

# Applied

## Q1

Rapid antigen tests are commonly used to detect **COVID-19 infection**. Suppose a new rapid test is compared against the gold standard PCR test.  

For this question, focus on **asymptomatic adults**. About **5% of asymptomatic adults** truly have COVID-19 infection. The test has a **sensitivity of 90%** and a **specificity of 95%** for detecting infection in this population.


a. Suppose you administer the rapid test to someone who truly has COVID-19. What is the probability that the test will return a positive result? Circle one.

(A) 5%  
(B) 90%  
(C) 10%  
(D) 95%  
(E) Not enough information to determine this probability  

**(B).** The probability that the test is positive given that the person truly has COVID-19 is just the **sensitivity**, 90% (0.9).

b. Let $A$ be testing positive and $B$ be having COVID-19. How would you calculate the probability of someone testing positive AND having COVID-19? You can leave your answer in terms of symbols.

$$
P(A \cap B) = P(B\vert A)P(B)
$$

\vskip 5cm

## Q2

A new rapid test has been developed to detect **hepatitis B infection**. A study compares the results of the rapid test against the gold standard laboratory test. The results from 200 patients are summarized below:

|                       | Gold Standard Positive | Gold Standard Negative | Total |
|-----------------------|----------------------|----------------------|-------|
| **Rapid Test Positive** | 70                   | 25                   | 95    |
| **Rapid Test Negative** | 10                   | 95                   | 105   |
| **Total**             | 80                   | 120                  | 200   |

**Question:**  
Calculate the **sensitivity** of the rapid test. Leave your answer as a fraction.

**From the table:**

**-   Test positive & disease positive = 70**

**-   Disease positive total = 80**

**$\text{Sensitivity} = \frac{70}{80}$**

## Q3

A researcher is studying the number of patients arriving at a hospital emergency room with asthma attacks during a 1-hour period. She proposes using a Poisson distribution to model the number of arrivals per hour.

Describe how you would verify whether each of the assumptions for a Poisson distribution are satisfied in this context. (e.g. what would you need to know/verify for each assumption, specific to this context?)

1.  **Events occur independently** – Each patient’s arrival is assumed independent of others.

    -   In practice, independence may be reasonable, unless e.g., an asthma trigger affects multiple people simultaneously.

2.  **The average rate** $\lambda$ is constant – The expected number of arrivals per hour does not change significantly over time.

3.  **Events occur singly** – Only one event occurs at a particular instant.

4.  **Number of events in non-overlapping intervals is independent** – The number of arrivals in one hour is independent of arrivals in another hour.

\vskip 3cm

## Q4

a. An environmental agency classifies daily air quality in a city into three categories: Good, Moderate, and Unhealthy. The probabilities of each category on a randomly selected day are summarized in the table below:

| Air Quality | Probability |
|-------------|------------|
| Good        | 0.5        |
| Moderate    | 0.3        |
| Unhealthy   | 0.2        |


State the Kolmogorov axioms. Are all axioms true for this example? 

All probabilities are between 0 and 1.

The sum of probabilities for all outcomes is 1.

The probability of the union of disjoint events equals the sum of their probabilities.

\vskip 3cm

## Q5

a. A public health researcher is studying the number of **asthma attacks requiring emergency care per day** in a mid-sized city. She models the number of attacks per day using a Poisson distribution with rate parameter $\lambda = 3$.  

Recall the Poisson PMF:  

$$P(X = k) = \frac{e^{-\lambda} \lambda^k}{k!}, \quad k = 0, 1, 2, \dots$$


Write out, in terms of the Poisson PMF, the probability that on a randomly selected day there are **at least 2 asthma attacks**. You do not need to compute the numeric value.  

$P(X \ge 2) = 1 - P(X < 2) = 1 - P(X = 0) - P(X = 1)$

$P(X \ge 2) = 1 - \frac{e^{-3} 3^0}{0!} - \frac{e^{-3} 3^1}{1!}$

b. Continuing part (a) above, Which of the following R commands correctly calculates the probability that on a randomly selected day there are **at least 2 asthma attacks**?

(A) `dbinom(1, lambda = 3)`  

(B) `1 - ppois(1, lambda = 3)`  

(C) `ppois(2, lambda = 3)`  

(D) `1 - dpois(2, lambda = 3)` 

**Answer is (B): `1 - dpois(0, lambda = 3) - dpois(1, lambda = 3)` note this is the same as `1 - ppois(1, lambda = 3)` since when we use ppois, vs dpois, this captures cumulative probability so the probability of X$\leq$ 1. But since we did not go over the difference between ppois or pbinom in class, I will only give you functions with dbinom or dpois - you are welcome though to use both**

## Code

Elena Epidata is an epidemiologist exploring health trends in the U.S. population. She is using a dataset similar to **NHANES**, which includes variables such as **age, BMI, blood pressure, and cholesterol** for a sample of adults. Elena is new to R and is practicing basic data exploration.

## Q6

Elena types the following code in R:

```{r}
#| echo: true
#| eval: false
library(tidyverse)
nhanes <- read_csv("nhanes_sample.csv")
View(nhanes)
glimpse(nhanes)
names(nhanes)
```

Which of the following statements are true about the outputs of these commands? Select all that apply.

(A) `View()` opens a spreadsheet-style view of the dataset.
(B) `glimpse()` prints a concise summary of the dataset to the console.
(C) `names()` lists the names of the columns.
(D) `View()` returns a vector of column names.

**A, B, C are correct.**


## Q7

Elena wants to examine a subset of the data. She types:

```{r}
#| echo: true
#| eval: false
nhanes_highbp <- nhanes |>
  filter(age >= 50, sbp > 140)
```

a. What subset of data has she created?

**those with age at least 50 and sbp greater than 140 mmHg**

\vskip 1cm

b. What does `|>` do?

**pipes the dataset into the filter command**

\vskip 1cm

c. Will anything be printed out by running this code?

**No. She is creating a new dataset but hasn't asked anything to be printed.**

\vskip 1cm

## Q8

Elena wants to create a scatter plot of BMI vs. systolic blood pressure and color points by sex. She writes:

```{r}
#| echo: true
#| eval: false
ggplot(nhanes, aes(x = bmi, y = sbp, color = sex)) +
  geom_point() +
  labs(x = "BMI", y = "Systolic Blood Pressure (mmHg)")
```

Which of the following statements are true? Select all that apply.

(A) `geom_point()` creates a scatter plot.
(B) Points are colored based on BMI.
(C) This code produces a boxplot of blood pressure.
(D) The x-axis represents BMI.

**A and D are correct.**

## Q9

Elena wants to calculate the average BMI by `age_group`. This variable has 3 categories: `young`, `mid`, and `older`. She types:

```{r}
#| echo: true
#| eval: false
nhanes_summary <- nhanes |>
  group_by(age_group) |>
  summarize(mean_bmi = mean(bmi, na.rm = TRUE))
```

What does `nhanes_summary` contain?

**One row per age_group and a column, mean_bmi, containing the mean BMI for each age group, with NAs removed.**

\vskip 2cm

## Conceptual (True/False)

Circle either True or False. 

## Q10

The Poisson distribution is often used to model counts of rare events occurring independently in a fixed time or space interval.

**True**

False 

## Q11

In a discrete probability distribution, all individual probabilities must be greater than 0.

True  

**False**  

## Q12

Bayes’ Rule allows you to compute the probability of an event given some observed evidence: P(A|B) = [P(B|A) * P(A)] / P(B).

**True**  

False  

## Q13

A frequentist interprets probability as a personal degree of belief, while a Bayesian interprets probability as the long-run relative frequency of events.

True  

**False**