---
title: "Exam 1 In-Class Review"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Format of the Exam

There will be three sections on Exam 01: Code, Conceptual (which will include some True/False), and Applied. 

We will practice questions from each of these sections as a group today, as well as go over any questions from past homeworks/Lab 1/the other review questions.

# Code

## Q1

Dr. X is a clinician learning R in order to perform some statistical analysis on her dataset. You are a biostatistician also working on the project. She's begun to learn the basics but has some questions for you.

What's the difference between `install.packages()` and `library()` ?

## Q2

Dr. X is tempted to simply Google some excel functions and analyze her data in excel, because R is difficult to learn. She wants to simply email her collaborators her final figures she creates from this process. Why is doing the analysis in excel problematic in terms of reproducibility? How can R help with reproducibility in Dr. X's analysis?

\vskip 5cm

## Q3

In a Quarto document, what does \### in front of text do?

\vskip 3cm

## Q4

In a code chunk within a Quarto document, what does \# in front of code do?

\vskip 3cm

## Q5

Dr. X is ready to begin her analysis and attempts to load in the tidyverse package but is getting an error. Based on her code below, why is she getting an error and how should she fix it?

```{r}
#| eval: false
#| echo: true
install.packages("tidyverse"))
```

\vskip 3cm

## Q6

Dr. X figured out her error above. She is now doing some practice visualization on the CDC dataset (the same one we worked on in Lab 2). She has produced the following code:

```{r}
#| eval: false
#| echo: true
ggplot(data = cdc, mapping = aes(x = Exercise, 
                                 y = Obesity, 
                                 color = HDI, 
                                 shape = HDI)) +
geom_point()
```

What does `geom_point()` do? (i.e. what kind of plot will it produce?)

\vskip 3cm

## Q7

Dr. X is producing a summary of some of the CDC variables below.

```{r}
#| eval: false
#| echo: true
cdc |>
  summarize(avg_exercise = mean(Exercise, na.rm = T))
```

a.  What is the `summarize()` function doing?

\vskip 3cm

b.  What does `na.rm = T` mean?

\vskip 3cm

## Q8

A new vaccine is tested, and researchers know that it successfully prevents infection in **80%** of people who receive it. Suppose 10 people are vaccinated. What R code could you use to calculate the probability that **exactly 7** of the 10 vaccinated individuals are protected from infection?

```         
A.  `dbinom(7, size = 10, prob = 0.4)`

B. `dbinom(10, size = 7, prob = 0.8)`  

C. `dbinom(7, size = 10, prob = 0.8)`  

D. `dpois(7, lambda = 8)` 
```


# Conceptual

## Q9

What's the difference between reproducibility and replicability?

\vskip 3cm

## Q10

What is quota sampling? Would this type of sampling be more or less generalizable to the broader population compared to probability sampling?

\vskip 3cm

## Q11

In class, we discussed the response to a COVID-19 seroprevalence paper, and how the investigators recruited their participants using Facebook. Describe why this method may have introduced a form of sampling bias.

\vskip 3cm

## Q12

Consider the following research question:

-   "Is the PCV13 vaccine effective against community acquired pneumonia in adults aged 65 or older?"

What is the study population?

\vskip 3cm


## Q13

True or False: Events $A$ and $B$ are said to be independent when $P(A \cap B) = P(A) \cdot P(B)$.


# Applied

## Q14

According to CDC growth charts, the mean systolic blood pressure for 60-year-old men is 135 mmHg, with a standard deviation of 20 mmHg. Suppose you are a 60-year-old man with a systolic blood pressure of 165 mmHg.

a.  What expression would you use to calculate your z-score?

\vskip 3cm

b.  Why are z-scores helpful when comparing normally distributed variables with different means and variances?

\vskip 3cm

## Q15 {.smaller}

A new rapid test has been developed to detect **influenza infection**. A study compares the results of the rapid test against the gold standard laboratory PCR test. The results from 200 patients are shown below:

|                         | PCR Positive | PCR Negative | Total |
|-------------------------|--------------|--------------|-------|
| **Rapid Test Positive** | 72           | 18           | 90    |
| **Rapid Test Negative** | 8            | 102          | 110   |
| **Total**               | 80           | 120          | 200   |

a.  Calculate the **sensitivity** of the rapid test. 

b.  Calculate the **specificity** of the rapid test. 

c.  In one sentence, explain what specificity means in this context.


