---
title: "Lab 03: Wrangling data with tidyverse"
author: "Your Name Here"
date: "Today's Date Here"
format: html
---

```{r }
#| label: setup
#| message: false
#| warning: false

library(tidyverse)

ncbikecrash <- read_csv("https://karamccor.github.io/b6002/labs/data/bikecrash.csv")
```

### Exercise 1

#### (a)

```{r ex-1}
#| label: ex-1a
#| eval: false

# run this code
ncbikecrash |> 
  filter(development == "Residential") |>
  filter(driver_age_group == "0-19")
```

[type response here]

#### (b)

```{r}
#| label: ex-1b

# edit code below: replace blank with Residential, then run code

ncbikecrash |> 
  filter(development == "______") |>
  count(driver_age_group)
```

### Exercise 2

```{r }
#| label: ex-2

# edit code: replace blank with driver estimated speed variable
ncbikecrash |> 
  count(____________) |>
  arrange(n)
```

### Exercise 3


```{r }
#| label: ex-3

# type code here
```

### Exercise 4

```{r }
#| label: ex-4

# type code here
```

### Exercise 5

```{r}
#| label: ex-5a

# type code here
```


```{r}
#| label: ex-5b

# type code here
```

### Exercise 6

```{r }
#| label: ex-6

# type code here
```

### Exercise 7

```{r}
#| label: ex-7
#| warning: false

# type code here
```

[write response here]

```{r}
#| label: ex-7b
#| warning: false
ggplot(ncbikecrash, mapping = aes(x = rural_urban, 
                                  y = driver_age, 
                                  fill = weather)) +
  geom_boxplot()
```


[write response here]

**Submission:** Remember to render your lab into an .html file and then save it as a .pdf in a web browser. You must turn in a .pdf file corresponding to the Quarto template to Gradescope in order to receive credit for the labs and homework assignments. You must assign each problem to a page in your PDF. Assign the Formatting portion to your page 1. Ensure all code is visible by indenting after commas as needed.  
