---
title: "Lab 6: Code Style"
format: html
---

Due Friday, March 6 at 11:59pm.

## Today's lab

Today's lab will focus on writing well-formatted and well-styled code.

Note: While we typically allow AI use for coding help, this assignment is particularly geared towards learning proper coding style yourself. Thus, **AI use is not permitted on this assignment**.

## Why do we care?

To quote Hadley Wickham,

-   Good coding style is like correct punctuation: you can manage without it, butitsuremakesthingseasiertoread.

Many of the conventions used here are made in order to increase readability of your code both for yourself and others. Additionally, keeping a consistent style throughout your code helps readability and makes it easier on you – once you’ve gotten accustomed to appropriate style choices, there are fewer decisions to make when writing code.

## Naming things

-   Use only lowercase letters, numbers (if needed), and underscores (`_`).

-   Underscores should be reserved for separating words within a name.

-   Variable names should be nouns and function names should be verbs.

-   Try your best to have concise and meaningful names (hard, at times!)

-   Where possible, avoid re-using names of common functions and variables in R.

## Spacing

-   Always put a space after a comma, and never before (just like in English).

-   Do not put spaces immediately inside or outside parentheses or brackets.

-   `==`, `+`, `-`, `*`, `/`, `~`, and `<-` should be surrounded by spaces.

-   `!`, `$`, `?`, `:`, and \^ should NOT be surrounded by spaces.

-   Extra spaces is ok if it improves alignment of functions. For instance:

```{r echo = T, eval = F}
# Good
height <- (feet * 12) + inches
mean(x, na.rm = TRUE)

sqrt(x^2 + y^2)
df$z
x <- 1:10

# Bad
height<-feet*12+inches
mean( x , na.rm=TRUE)

sqrt(x ^ 2 + y ^ 2)
df $ z
x <- 1 : 10
```

## Quotes

Use `" "`, not `' '`, for quoting text. The only exception is when the text already contains double quotes and no single quotes.

```{r echo = T, eval = F}
ggplot(diamonds, mapping = aes(x = price)) +
  geom_histogram() +
  labs(title = "Shine bright like a diamond", # Good
       x = "Diamond prices",                  # Good
       y = 'Frequency')                       # Bad
```

## Curly braces

Curly braces help define R code hierarchy, for instance in defining functions and if-else statements. A few rules should be kept in mind when using curly braces:

-   `{` should be the last character on the line. Related code (e.g., an if clause, a function declaration, a trailing comma, etc.) must be on the same line as the opening brace.

-   The contents should be indented by two spaces.

-   `}` should be the first character on the line.

Note: We have not talked about functions like the one below, so don't worry if they seem confusing. We will give a high-level overview, but no need to get bogged down with the function content. The purpose of this example is just to see proper curly brace alignment.

```{r echo = T, eval = F}
# Good
if (y < 0 && debug) {
  message("y is negative")
}

if (y == 0) {
  if (x > 0) {
    log(x)
  } else {
    message("x is negative or zero")
  }
} else {
  y^x
}

# Bad
if (y < 0 && debug) {
message("Y is negative")
}

if (y == 0)
{
    if (x > 0) {
      log(x)
    } else {
  message("x is negative or zero")
    }
} else { y ^ x }
```

## Long lines

Try to keep your code and narratives to 80 characters per line, which fits comfortably on a printed page with reasonably sized font. In order to get a guide at the 80-character limit, you may go to the following menu in your R menu bar:

Tools \> Global Options \> Code \> Display (tab) \> Show margin at 80 characters.

## Pipes for `dplyr`

-   Pipes should always have spaces before it.

-   Pipes should always be followed by a new line.

-   After the first step, each line should be indented by two spaces.

```{r echo = T, eval = F}
# Good
iris |>
  group_by(Species) |>
  summarize_if(is.numeric, mean) |>
  ungroup() |>
  gather(measure, value, -Species) |>
  arrange(value)


# Bad
iris |> group_by(Species) |> summarize_all(mean) |>
ungroup |> gather(measure, value, -Species) |>
arrange(value)
```

## Layering in `ggplot2`

-   `+` for new layers should always have spaces before it.

-   `+` should always be followed by a new line.

-   After the first step, each line should be indented by two spaces.

## Long lines

If arguments to a function don’t all fit on a single line (for instance, when creating axis labels and titles), put each argument on its own line and indent to line up with the hierarchy.

```{r echo = T, eval = F}
# Good
ggplot(aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
  geom_point() +
  labs(
    x = "Sepal width, in cm",
    y = "Sepal length, in cm",
    title = "Sepal length vs. width of irises"
  ) 

# Bad
ggplot(aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
  geom_point() +
  labs(x = "Sepal width, in cm", y = "Sepal length, in cm", title = "Sepal length vs. width of irises") 
```


## Your turn!

In each of the exercises in the lab template, fix each code and/or narrative example given in the template to conform to good R style guidelines as denoted above. 

**Do not actually run the code. You will get an error if you try to run the code, as objects are not defined.**

For each exercise, you should check the following, if applicable:

- Naming things - do names follow good naming convention?

- Spacing

- Quotes

- Curly braces

- Long lines

- Pipes for `dplyr`

- Layering in `ggplot`

- Long lines

The examples have been pre-loaded in the templates for your convenience. The template is available in the Canvas assignment page.

It is not important for you to understand what the code does. Do not evaluate any of the code (i.e., keep `#| eval: FALSE` in the R chunks).

## Acknowledgements

Today’s guide follows the tidyverse style guide by Hadley Wickham and extensively quotes his guide. Credit should be due exclusively to him for the guidelines and examples.

A more complete version of the style guide may be found [here](https://style.tidyverse.org/).


## Submission

As you’ve seen previously, we can **Render** the template into an .html file that can be opened by any web browser. To export it as a .pdf, open the file in your web browser and then print to or save as a .pdf document. Your TAs will show you how if you need help! (There is a way to directly knit to a .pdf file, but it’s quite a bit more involved.)

You will submit the PDF documents for labs and homework to Gradescope as part of your final submission.

To submit your assignment:

- Access Gradescope through the menu on the BIOS 600 Canvas site.

- Click on the assignment, and you’ll be prompted to submit it.

- Mark the pages associated with each exercise. All of the pages of your lab should be associated with at least one question (i.e., should be “checked”).

- Select the first page of your .PDF submission to be associated with the “Formatting” section.


## Grading

| Component | Points |
|----------|--------|
| Ex 1 | 4 |
| Ex 2 | 4 |
| Ex 3 | 4 |
| Ex 4 | 4 |
| Ex 5 | 4 |
| AI Attestation| 1 |
| Formatting | 3 |

The “Formatting” grade is to assess the document format. This includes having a neatly organized document (no excessive output, warnings/messages when loading packages and/or data) with readable code and your name and the date updated in the YAML.
