---
title: "Practice Problems on Collinearity and Logistic Regression"
format: html
---

Some of these problems are meant to be fairly difficult--the purpose is to get you to think carefully about the concepts we've covered in class, and give you a better understanding of those underlying concepts, in order to be better prepared for the final.

Since some of these are a bit challenging, you are encouraged to discuss these problems in small groups.

## Exercise 1: Multicollinearity and linear regression

A (fictional) study examined the association between three air pollutants, Pollutant1, Pollutant2, and Pollutant3, and prevalence of asthma within counties in a fictional U.S. state. The pollutant variables are the average concentrations in parts per billion (ppb) measured at monitoring sites within that county. Prevalence of asthma is the % of the county's population diagnosed with asthma; it's treated as normally distributed. The sample comprised 100 counties in this state. Assume hypothesis tests are conducted with significance level $\alpha=.05.$

```{r}
#| echo: false

load("collinear_logistic_practice_data.Rdata")
fit <- lm(asthma ~ ., data=df)
summary(fit)

```

### (a)

What are the results of the hypothesis tests for the significance of each pollutant? Interpret these results in the context of the study.

### (b)

Write the null and alternative hypotheses for the F-test for overall model significance. Write these hypotheses both in symbols, and in words in the context of the study.

### (c)

What is the result of the overall F-test for the linear regression model? Interpret this result in the context of the study.

### (d)

Do these results show indication of possible multicollinearity affecting this model? Why or why not?

### (e)

The researchers decide to check for multicollinearity as follows:

```{r}
library(car)
vif(fit)

```

What did the researchers do? Explain in 1 sentence. What do the results indicate about multicollinearity affecting this model?

## Exercise 2: Conceptual

Assume the outcome variable $Y$ takes values either $0$ or $1$.

### (a)

If our outcome variable $Y$ takes values $0$ or $1$, what would we assume its distribution is?

### (b)

We usually want to create a regression model to estimate $P(Y_i=1)$ based on the value of the predictors for observation $i$. Thinking back to the properties of the distribution you said in question (a), what is another name for this quantity? In other words, how else can we think of $P(Y_i=1)$?

### (c)

Based on your answer to part (c), how is this similar to the linear regression model with a normally-distributed outcome?

### (d)

Why would or wouldn't you want to use a linear regression model for this situation?

### (e)

What is the logistic function? Why do we use it in logistic regression?

## Exercise 3: Logistic Regression

A (fictional) study examined the association between age (years) and region of residence in the U.S. with diagnosis of a particular condition. They used four regions of the continental U.S., values 1-4.

```{r}
#| echo:  true

fit.glm <- glm(condition ~ ., family="binomial", data=df2)
summary(fit.glm)

```

### (a)

What is the reference category for the region variable?

### (b)

Write the model used.

### (c)

Conduct a hypothesis test for the association between age and diagnosis of the condition, at level $\alpha=.05$. Follow the steps:

1.  Write the null and alternative hypotheses, in words **and** symbols.
2.  Compare the p-value to $\alpha$.
3.  Draw and state your conclusion in the context of the study.

### (d)

Interpret the coefficient for the age variable in the context of the study.

### (e)

Interpret the coefficient for region category 4 in the context of the study.

### (f)

Based on the sign (positive/negative) of $\hat{\beta}_1$, is increased age associated with increased or decreased probability of diagnosis?
